::: {.content-hidden}
{{< include ../latex-header.tex >}}
:::

# Drawing from Densities {#sec-draws}

<!-- Sources: DCMS Ch 9 (lit/notes/train-2009-dcms.md); Rchoice §2.3/§4 for draw-transformation formulas (lit/notes/pkg-rchoice.md); Rizzo for statistical computing background -->

## Why We Need Draws

<!-- draws appear three times in this book: simulating data from a model (ch 3), approximating integrals in MSL (ch 12), and exploring posteriors via MCMC (ch 15-17); everything reduces to uniform draws plus transformations -->

## Pseudo-Random Numbers from the Uniform Distribution

<!-- what a PRNG is (briefly); set.seed() and reproducibility; runif() -->

## The Inverse Transform Method

$$
\begin{aligned}
F_X(x)
    &= \Pr \left(X \le x \right) \\
    &= \Pr \left(T(U) \le x \right) \\
    &= \Pr \left(U \le T^{-1}(x) \right) \\
    &= T^{-1}(x)
\end{aligned}
$$

### Example: Exponential

$$ F(y,\lambda) = 1 - \exp\left( -\lambda y \right) \hspace{1em} \text{for} \ y \ge 0 $$

$$ F^{-1}(u) = -\log(1-u) / \lambda $$

```{r}
n <- 1e4
lambda <- 0.3
u <- runif(n)
y <- -log(1-u)/lambda
```

```{r}
x <- seq(from=0, to=ceiling(max(y)), length.out=100)
hist(y, freq=F, breaks=100)
lines(x=x, y=dexp(x,lambda), lwd=2)
```

### Example: Gumbel

$$ F(y) = \exp \left( -\exp \left( -y \right) \right) $$

$$ F^{-1}(u) = -\log(-log(u))  $$

```{r}
n <- 1e4
u <- runif(n)
y <- -log(-log(u))
```

```{r}
x <- seq(from=-3, to=ceiling(max(y)), length.out=500)
hist(y, freq=F, breaks=100)
lines(x=x, y=ordinal::dgumbel(x), lwd=2)
```

### Example: Logistic

$$ F(y) = \frac{1}{1+\exp \left( -x \right)} $$

$$ F^{-1}(u) = \log(u) - \log(1-u) $$

```{r}
n <- 1e4
u <- runif(n)
y <- log(u) - log(1-u)
```

```{r}
x <- seq(from=floor(min(y)), to=ceiling(max(y)), length.out=500)
hist(y, freq=F, breaks=100)
lines(x=x, y=dlogis(x), lwd=2)
```

### Example: Normal

```{r}
n <- 1e4
y <- rnorm(n)

u <- runif(n)
y <- qnorm(u)
```

## Accept-Reject Methods

<!-- for densities without invertible CDFs; simple example; note MCMC (ch 15) as the eventual general-purpose answer -->

## Transforming Draws

<!-- location-scale: mu + sigma*z; lognormal via exp(z) (preview: price coefficients in ch 13); triangular/uniform via inverse transform; see Rchoice §2.3 for the standard menu -->

## The Multivariate Normal

<!-- Cholesky root: y = mu + L z; verify with empirical covariance; this exact operation reappears in ch 13 (correlated random coefficients) and ch 16-17 (hierarchical draws) -->

### Example: Correlated Normal Draws

## Truncated Distributions

<!-- truncated normal via inverse-CDF on a restricted range; needed for data augmentation in ch 16 (Albert-Chib) and ch 19-20 (MNP Gibbs) -->

## Quasi-Random Draws

<!-- Halton sequences and antithetic draws; why coverage beats randomness for integration; DCMS Ch 9; used in ch 12-13 (MSL) -->

## Key Learnings
