::: {.content-hidden}
{{< include ../latex-header.tex >}}
:::

# Organizing Choice Data {#sec-data}

```{r}
#| label: setup
#| message: false
#| echo: false
library(ggplot2)
theme_set(theme_minimal())
```

This chapter contains no new statistics. It is about something more mundane and, in my experience, responsible for more student-hours of frustration than any mathematical topic in this book: getting choice data into a shape that estimation code can consume. The likelihoods of Part II are sums over choice situations of expressions involving each alternative's attributes; between the data as they arrive and the likelihood as it computes sits a layer of pure bookkeeping. Do the bookkeeping haphazardly and every later chapter fights it. Do it once, deliberately, and the same few structures will carry us from the plain MNL (@sec-mnl-mle) through simulated likelihood (@sec-mixed-msl) to the hierarchical samplers (@sec-hb-mnl) without modification.

Because this is a chapter of choices-with-tradeoffs rather than theorems, I will follow a pattern used throughout the book: lay out the reasonable options, then commit to one and say why. You may prefer different choices in your own work --- the options all *work* --- but a book, like a codebase, benefits from picking a lane.

## The Shape of Choice Data

The data of a regression course are flat: one row per observation, one column per variable. Choice data have more structure, because each observation is not a number but an *event with internal anatomy*: a particular person, facing a particular set of alternatives, on a particular occasion, choosing one. Three indices track that anatomy, and they have been fixed since @sec-simdata:

- $n = 1, \ldots, N$ indexes **decision-makers** (our survey respondents),
- $t = 1, \ldots, T$ indexes **choice situations** (the tasks each respondent completes),
- $j = 1, \ldots, J$ indexes **alternatives** within a choice situation.

The pair $(n, t)$ identifies one choice situation; the triple $(n, t, j)$ identifies one alternative in one situation --- one *utility* in the model's terms, since @sec-mnl will assign $U_{ntj} = \x_{ntj}'\bfbeta + \varepsilon_{ntj}$ to exactly this triple. Our laptop study (@sec-running_example) has $N = 500$, $T = 8$, and $J = 3$: five hundred respondents, eight tasks each, three laptop profiles per task, for $500 \times 8 = 4{,}000$ observed choices and $4{,}000 \times 3 = 12{,}000$ described alternatives.

The binary data of @sec-simdata were the degenerate case that hid all of this: one task per person ($T=1$) and one modeled alternative plus a fixed outside option ($J=2$ with the "don't buy" utility normalized away), so a flat data frame sufficed. It no longer does. The question of this chapter is: *in what shape do we store 12,000 alternative descriptions and 4,000 choices?*

## Wide and Long Formats

Two conventions dominate, distinguished by what a row represents.

In **wide format**, one row is one choice situation $(n,t)$, and each alternative's attributes spread across columns:

```{r}
#| echo: false
wide_ex <- data.frame(
  id = c(1L, 1L), task = c(1L, 2L),
  brand_1 = c("Acer","Dell"),  price_1 = c(1.2, 1.6),
  brand_2 = c("Apple","Acer"), price_2 = c(2.0, 0.9),
  brand_3 = c("Dell","Apple"), price_3 = c(1.4, 1.8),
  choice = c(2L, 1L)
)
wide_ex
```

In **long format**, one row is one alternative $(n,t,j)$, with the choice recorded as an indicator:

```{r}
#| echo: false
long_ex <- data.frame(
  id     = rep(1L, 6),
  task   = rep(c(1L, 2L), each = 3),
  alt    = rep(1:3, times = 2),
  brand  = c("Acer","Apple","Dell","Dell","Acer","Apple"),
  price  = c(1.2, 2.0, 1.4, 1.6, 0.9, 1.8),
  choice = c(0L, 1L, 0L, 1L, 0L, 0L)
)
long_ex
```

(These two tables show the same two tasks; respondent 1 chose the Apple in task 1 and the Dell in task 2.)

The tradeoffs:

- **Wide** matches how survey platforms export data and how humans read a choice task ("here were your three options; you picked #2"). It is compact --- one row per choice. But the attributes of "alternative 2" live in differently-named columns from "alternative 1," so any computation over alternatives means wrangling column-name patterns; and if some tasks show two alternatives while others show four, the format needs padding columns and missing-value conventions.
- **Long** matches how the *model* thinks. Each row is one $(n,t,j)$ triple --- one utility. Attributes have a single column each regardless of $J$; tasks with different numbers of alternatives are just groups of different sizes; and the operations our likelihoods need (compute a value for every alternative, then aggregate within choice situations) become operations on rows and groups.

**Our choice: long format, always.** The deciding argument is the last one --- in @sec-mnl-mle the log-likelihood will be built from exactly two moves, "one value per row" and "aggregate rows within task," and long format makes both natural. Wide-to-long conversion is a mechanical reshape,[^1] so nothing is lost by standardizing. When a dataset in this book is called *the data*, it is a long-format data frame with columns `id`, `task`, `alt`, `choice`, then attributes.

[^1]: `stats::reshape()`, `tidyr::pivot_longer()`, or the `dfidx` package all automate it. We won't dwell on the mechanics; what matters is knowing which shape you're in and which shape you need.

One convention completes the format, and I state it as a contract because code will rely on it silently: **rows are sorted by `id`, then `task`, then `alt`.** Grouped computations in the chapters ahead identify a choice situation by contiguous runs of rows; the sort order is what makes that identification valid. Violate the contract and nothing errors --- results are simply, quietly wrong. (We will build one guardrail below.)

## Two Kinds of Variables, Two Kinds of Columns

Look again at the long-format table and note what varies where. `price` varies across rows *within* a task --- each alternative has its own. But a respondent characteristic like age would repeat across all rows of a task, varying only with `id`. The distinction matters enough for names:

- **Alternative-specific variables** ($\x_{ntj}$): attributes of the alternatives --- price, brand, memory. These vary within a choice situation and are the natural drivers of choice.
- **Individual-specific variables** ($\mathbf{z}_n$): characteristics of the decision-maker --- age, income, prior brand ownership. Constant across the alternatives in a task. (A notation caution: this $\mathbf{z}_n$ --- observed characteristics --- is unrelated to the standard-normal draw vector $\mathbf{z}$ of @sec-mvnorm; the collision is regrettably standard in this literature, and context always disambiguates.)

Here is the fact that surprises newcomers, stated now and explained properly in @sec-mnl: *an individual-specific variable cannot enter utility with a single common coefficient.* If age adds $\gamma \cdot \texttt{age}_n$ to the utility of **every** alternative, it raises all $J$ utilities equally, changes no comparison, and drops out of the choice probabilities entirely --- $\gamma$ is unidentified. Only *differences* in utility move choices. Individual-specific variables earn their way into choice models by interaction: with alternative dummies (age shifts the appeal of Apple relative to Acer) or with attributes (income damps price sensitivity). Structurally, that means $\mathbf{z}_n$ columns in a long data frame are raw material awaiting interaction, not regressors in their own right. In this book's main thread, individual-specific variation instead enters through *coefficients that vary across people* --- the heterogeneity models of @sec-latentclass onward --- and in @sec-hb-mnl the two devices meet, with $\mathbf{z}_n$ shifting the *mean* of person $n$'s coefficient vector.

## Coding Categorical Attributes

Brand takes values Acer, Dell, Apple; memory takes 8, 16, 32. Utilities are sums of numbers, so categories must become numeric columns, and the mapping is another small decision with options.

**Dummy (treatment) coding** picks a reference level and creates an indicator per remaining level. With Acer as reference: `dell` and `apple` columns, each 0/1. A brand coefficient is then the utility of that brand *relative to Acer, all else equal* --- concrete and readable, with the reference level absorbing the baseline.

**Effects coding** constrains the level utilities to sum to zero instead: the reference level's columns are all $-1$, and each level's utility is read relative to the *average* level. Conjoint practice often prefers it because part-worths then compare against a neutral average rather than an arbitrary reference, and it interacts more gracefully with the "attribute importance" summaries of @sec-postestimation [see @eggers2022 for conventions in applied CBC work].

**Our choice: dummy coding, with the reference levels fixed once** --- Acer, 8GB, 13-inch, as set in @tbl-laptop-attributes. For a book, interpretability-per-line-of-code wins: "the Apple coefficient is the extra utility of an Apple over an Acer" needs no further explanation, and every simulate-and-recover check reads directly against true parameters stated in the same coding. Effects coding returns for a cameo in @sec-postestimation where its advantages actually bite. What is *not* optional is consistency: coding schemes, like sort orders, are contracts, and estimates are only interpretable against the coding that produced them.

R can build coded columns from factors automatically via `model.matrix()`, and we will let it --- with the factor levels ordered explicitly so the reference is what we intend, not what alphabetical order happens to deliver:

```{r}
brand <- factor(c("Acer","Apple","Dell"), levels = c("Acer","Dell","Apple"))
model.matrix(~ brand)
```

Note `model.matrix()` includes an intercept column by default. Whether we keep it is the next section's business.

## Where Is the Intercept? {#sec-data-intercept}

The binary model of @sec-simdata had an intercept: a constant in $\x_n$ giving the baseline pull toward buying. Scan the long-format laptop data for its analogue and you hit a small conceptual wall: a column of ones would give *every alternative in every task* the same utility boost --- and as with individual-specific variables, anything common to all alternatives cancels from every comparison. A common intercept is unidentified in multinomial choice data.

What can exist is an **alternative-specific constant (ASC)**: a dummy for *position* $j$, capturing systematic utility for "option 2" as such. Whether ASCs belong in the model depends on the design:

- With **labeled alternatives** --- the options are inherently "car," "bus," "train," the staples of transportation economics [@train2009, Section 2.5] --- ASCs are essential. Each mode has unmeasured systematic appeal, and its ASC absorbs it.
- With **unlabeled profiles** --- our study, and most conjoint: the three laptops in a task are just "some laptops," their screen positions randomized --- position has no meaning, and the brand attribute already carries the "which product" information. Including position ASCs would estimate, roughly, left-middle-right screen bias.[^2]

[^2]: Which is a real (if small) behavioral phenomenon, and estimating position effects is occasionally a useful design diagnostic. The point is that in an unlabeled design ASCs are a diagnostic, not a structural part of utility.

So our laptop data frames carry no intercept and no ASCs; the brand dummies play the role closest to "brand intercepts." When Part IV needs ASCs (the probit's identification story makes them central) and when @sec-extensions adds a no-choice option (whose ASC is the entire model of "walking away"), we will add the relevant dummy columns explicitly, as ordinary attributes. That is a quiet advantage of long format: an ASC is just one more 0/1 column, not a special modeling object.

## Building the Study Design {#sec-build-design}

Enough principles; let's build the actual dataset scaffold for the laptop study. A useful separation of concerns --- and the one real studies force on you anyway --- is between the **design** (which profiles appear in which tasks; exists before any respondent) and the **responses** (the `choice` column; exists after). We build the design now. Filling in choices requires the MNL model itself, so that half arrives in @sec-mnl.

How should profiles be assigned to tasks? The gold standard is a formal experimental design --- orthogonal arrays, D-efficient designs --- that maximizes the information the study extracts per task [@rao2014, Chs. 2 and 4]. Design theory is a book of its own and deliberately out of our scope. We use the simplest defensible method, **independent random assignment**: every attribute of every profile drawn independently at random. Random designs are what our simulate-and-recover studies need (no accidental confounding between attributes, by construction) and they let the code stay transparent:[^tnote]

```{r}
sim_laptop_design <- function(N, T, J) {
    n_rows <- N * T * J
    data.frame(
        id    = rep(seq_len(N), each = T * J),
        task  = rep(rep(seq_len(T), each = J), times = N),
        alt   = rep(seq_len(J), times = N * T),
        brand = factor(sample(c("Acer","Dell","Apple"), n_rows, replace = TRUE),
                       levels = c("Acer","Dell","Apple")),
        ram   = factor(sample(c("8","16","32"), n_rows, replace = TRUE),
                       levels = c("8","16","32")),
        screen = factor(sample(c("13","15"), n_rows, replace = TRUE),
                        levels = c("13","15")),
        price = round(runif(n_rows, min = 0.8, max = 2.4), 2)
    )
}

set.seed(40)
laptops <- sim_laptop_design(N = 500, T = 8, J = 3)
head(laptops, 6)
```

[^tnote]: A confession: `T` is also R's abbreviation for `TRUE`, and masking it is normally poor hygiene. We use it anyway so the code reads like the math ($N$, $T$, $J$). Never abbreviate `TRUE` as `T` in your own code --- this is exactly why.

The `rep()` calls in the first three columns implement the sort-order contract mechanically --- `id` slowest, `alt` fastest --- so the design is born correctly ordered. Dimensions check:

```{r}
nrow(laptops)                       # 500 * 8 * 3
```

And a design-level sanity check in the spirit of @sec-sanity-checks: with independent random assignment, attributes should be uncorrelated and levels balanced. Brand shares by alternative position:

```{r}
round(prop.table(table(laptops$brand, laptops$alt), margin = 2), 2)
```

Near-uniform everywhere, as designed.

## From Data Frame to Likelihood Inputs {#sec-choice-data}

The last mile. A data frame is for humans; estimation code wants numeric arrays. Specifically, everything from @sec-mnl-mle onward needs three things: the attribute matrix, the choices, and a map from rows to choice situations. There are three reasonable containers, and this is the load-bearing decision of the chapter:

- **(a) A list of per-task matrices.** `data[[i]]` holds task $i$'s $J \times K$ attribute matrix and its choice. Transparent --- each likelihood term has its own little bundle --- and the shape most textbook pseudo-code implies. But R-level loops over 4,000 list elements are slow (@sec-performance will quantify how slow), and slow inside a likelihood means slow *everywhere*, because the likelihood is the inner loop of every method in this book.
- **(b) A three-dimensional array.** Dimensions (task, alternative, attribute): `X[i, j, k]`. Fast, vectorizable, and appealing when every task shows exactly $J$ alternatives. But it hard-codes the balanced-design assumption --- add a no-choice alternative to some tasks (@sec-extensions) or work with real-world availability variation, and the array needs padding and masking. Array code also reads poorly; three-index gymnastics obscure the model.
- **(c) A stacked matrix plus index vectors.** Keep the long format's row layout: one $(N \cdot T \cdot J) \times K$ matrix `X` whose rows are alternatives in contract order, a 0/1 vector `choice` marking chosen rows, and an integer vector `task_id` assigning each row to its choice situation. Grouped operations --- "sum $e^{V}$ within each task" --- become single calls to R's grouped-aggregation primitives (`rowsum()` chief among them), which run at compiled speed. Ragged tasks cost nothing: a task is however many rows share its `task_id`.

**Our choice: (c), the stacked-matrix representation.** It is within a whisker of the array's speed, it degrades gracefully to ragged designs, and --- decisive for this book --- it is the representation that survives unchanged into later parts: the MSL of @sec-mixed-msl evaluates the same stacked `X` under many coefficient draws, and the person-by-person Metropolis steps of @sec-hb-mnl slice it by `id` without restructuring. Build once, reuse for nineteen chapters.

The conversion function:

```{r}
build_choice_data <- function(df, attr_formula) {
    stopifnot(!is.unsorted(order(df$id, df$task, df$alt)))
    X <- model.matrix(attr_formula, data = df)
    X <- X[, colnames(X) != "(Intercept)", drop = FALSE]
    task_id <- cumsum(!duplicated(df[, c("id", "task")]))
    list(
        X       = X,
        choice  = if ("choice" %in% names(df)) df$choice else NULL,
        task_id = task_id,
        id      = df$id,
        n_tasks = max(task_id),
        K       = ncol(X)
    )
}
```

Every line earns its place:

- The `stopifnot()` guardrail enforces the sort-order contract; better a loud failure now than silent nonsense later.
- `model.matrix(attr_formula, df)` expands factors into dummy columns using their declared reference levels --- the coding decision of this chapter, executed by base R. We then drop the intercept column it insists on adding, for the identification reason of @sec-data-intercept.
- `task_id` is built by a small idiom worth reading twice: `!duplicated()` over the `(id, task)` pairs flags each task's first row, and `cumsum()` turns those flags into a running task counter --- `1,1,1,2,2,2,...` --- giving every row its group label in two vectorized calls.
- `choice` is optional, so the same function serves a bare design (now) and responded data (from @sec-mnl on).

Applied to the laptop design:

```{r}
dat <- build_choice_data(laptops, ~ brand + ram + screen + price)
str(dat, max.level = 1)
colnames(dat$X)
```

There it is: a `r nrow(dat$X)` $\times$ `r dat$K` matrix whose six columns[^3] --- `dell`, `apple`, `ram16`, `ram32`, `screen15`, `price` in our shorthand --- line up exactly, column for column, with the true parameter vector we will fix when the full study's choices are simulated in @sec-mnl, plus the index vectors that give the rows their meaning. When @sec-mnl-mle computes its first multinomial log-likelihood, it will consume precisely this object, and so will nearly every estimator after it.

[^3]: `model.matrix()` names them `brandDell`, `brandApple`, `ram16`, `ram32`, `screen15`, `price`: each factor contributes levels-minus-one dummy columns, per the coding discussion above.

## Package Conventions: `dfidx` and Friends

When we validate our hand-coded estimators against R packages --- `mlogit` in @sec-mnl-mle, `gmnl` in @sec-latentclass --- we must feed those packages *their* expected structure. Happily, the ecosystem standardized on ideas congruent with ours: `mlogit` [@croissant2020] consumes data via the `dfidx` package, which wraps a long-format data frame with declared indexes --- `idx = c("chid", "alt")` naming the choice-situation and alternative columns, with a nesting structure for respondent panels. The conversion from our canonical format is one line, which we will write when we first need it. If you understand this chapter, no R choice package's data layer will ever mystify you: they are all long format plus index bookkeeping, differing only in surface conventions.

## Key Learnings

- Choice data live on three indices --- decision-maker $n$, choice situation $t$, alternative $j$ --- and one row of our canonical data is one $(n,t,j)$ triple: **long format**, sorted by `id`, `task`, `alt`. The sort order is a contract that grouped computations rely on silently.
- Only utility *differences* move choices. That single fact organized the whole chapter: individual-specific variables need interactions to matter, a common intercept is unidentified, and ASCs are position dummies that belong in labeled designs but not in our unlabeled profile study.
- Categorical attributes become dummy columns against fixed reference levels (Acer, 8GB, 13"); `model.matrix()` mechanizes the coding, and consistency is what keeps estimates interpretable.
- The study **design** exists before responses: `sim_laptop_design()` builds the $500 \times 8 \times 3$ laptop study with independent random attribute assignment, correctly ordered by construction.
- The load-bearing decision: estimation code consumes the **stacked-matrix representation** --- `X` ($NTJ \times K$), 0/1 `choice`, integer `task_id` --- chosen over per-task lists (slow) and 3-D arrays (rigid) because it is fast, ragged-friendly, and reusable verbatim from MLE through MSL to MCMC. `build_choice_data()` performs the conversion and guards the contracts.
- Package data layers (`dfidx` for `mlogit`) are the same ideas under different names; conversion is mechanical.
