::: {.content-hidden}
{{< include ../latex-header.tex >}}
:::

# Code Performance {#sec-performance}

```{r}
#| label: setup
#| message: false
#| echo: false
library(ggplot2)
theme_set(theme_minimal())
```

This chapter pauses the march of models to sharpen the tools. It exists because of a compounding arithmetic that is easy to state and easy to underestimate: *the log-likelihood function sits at the bottom of everything, and everything above it multiplies.* A slow likelihood is a minor annoyance in @sec-mnl-mle and a fatal flaw by @sec-hb-mnl. The good news: for the models in this book, a handful of transferable techniques --- measure first, vectorize the hot path, precompute the invariant, preallocate the storage --- buy factors of hundreds, and we already own most of them.

## Why Speed Matters Now

Count the likelihood evaluations implied by the chapters ahead. One BFGS fit of the MNL costs on the order of a hundred evaluations (iterations times gradient arithmetic). The replication study of @sec-mnl-recovery multiplied that by 200. And the road ahead is steeper:

| Consumer | Evaluations of the (log-)likelihood kernel |
|---|---:|
| One BFGS fit (@sec-mnl-mle) | $\sim 10^2$ |
| Replication study, 200 runs (@sec-mnl-recovery) | $\sim 10^4$ |
| One MSL fit, $R = 200$ draws (@sec-mixed-msl) | $\sim 10^4$--$10^5$ |
| One MCMC run, $10^4$ iterations, per-person steps (@sec-hb-mnl) | $\sim 10^6$ |
| Prior sensitivity: several MCMC runs (@sec-hb-mnl) | $\sim 10^7$ |

: What will call your likelihood, and how often. {#tbl-eval-count}

A likelihood evaluation costing one second makes the last row a four-month computation; at one millisecond, three hours. Every method in Parts II--IV was *designed* in an era when this arithmetic bound what research was possible, and it still bounds what your afternoon is possible. That is the entire motivation; now the craft.

## Measure First

The cardinal rule of performance work: **never optimize what you have not measured.** Intuition about where time goes is famously unreliable, and effort spent speeding up a line that consumes 2% of runtime is effort donated to entropy. R's built-in stopwatch is `system.time()`; for the quick comparisons this chapter makes, we wrap it to report a median across repetitions (timings are noisy --- garbage collection, OS scheduling --- and medians resist the noise)[^lazy]:

```{r}
time_ms <- function(expr, times = 5) {
    e  <- substitute(expr)
    pf <- parent.frame()
    t  <- replicate(times, system.time(eval(e, pf))[["elapsed"]])
    round(1000 * median(t), 1)
}
```

[^lazy]: A subtlety worth stealing: R evaluates function arguments lazily and caches the result, so a naive `replicate(times, system.time(expr))` would time the real computation once and a cached lookup thereafter --- the classic way to convince yourself your code is infinitely fast. `substitute()` captures the unevaluated expression so each repetition genuinely recomputes. Measurement code deserves the same skepticism as model code.

For *profiling* --- finding where a complex computation spends its time, function by function --- R provides `Rprof()` and the friendlier `profvis` package; they matter once code has many moving parts, and I commend them to you for your own projects. Our computations are small enough that targeted timing tells the story.

The test subject is the laptop study likelihood from @sec-mnl-mle, with its data rebuilt as usual:

```{r}
study   <- readRDS("../data/laptop_study.rds")
laptops <- study$data
X <- model.matrix(~ brand + ram + screen + price, data = laptops)[, -1]
y <- laptops$choice
task_id <- cumsum(!duplicated(laptops[, c("id", "task")]))
beta <- study$beta_true
```

## Vectorizing the MNL Log-Likelihood

@sec-mnl-mle wrote the likelihood twice and promised a measurement. Here are three implementations, spanning the design space:

**Version 1 --- the pure loop** (from @sec-mnl-mle): visit each task, locate its rows with `which(task_id == i)`, compute locally.

```{r}
loglik_v1 <- function(beta, y, X, task_id) {
    ll <- 0
    for (i in unique(task_id)) {
        rows <- which(task_id == i)
        V    <- X[rows, , drop = FALSE] %*% beta
        ll   <- ll + V[y[rows] == 1] - log(sum(exp(V)))
    }
    as.vector(ll)
}
```

**Version 2 --- the hybrid**: compute all utilities in one matrix product (vectorized), but handle the per-task aggregation by splitting into a list and looping:

```{r}
loglik_v2 <- function(beta, y, X, task_id) {
    V  <- as.vector(X %*% beta)
    Vs <- split(V, task_id)
    ys <- split(y, task_id)
    ll <- 0
    for (i in seq_along(Vs)) {
        ll <- ll + Vs[[i]][ys[[i]] == 1] - log(sum(exp(Vs[[i]])))
    }
    ll
}
```

**Version 3 --- fully grouped** (the @sec-mnl-mle production version): no explicit loop at all; `rowsum()` performs every task's aggregation inside compiled code.

```{r}
loglik_v3 <- function(beta, y, X, task_id) {
    V     <- as.vector(X %*% beta)
    Vmax  <- ave(V, task_id, FUN = max)
    eV    <- exp(V - Vmax)
    denom <- rowsum(eV, task_id)
    sum( (V - Vmax - log(denom[task_id]))[y == 1] )
}
```

Same mathematics, same output (check it --- always):

```{r}
c(v1 = loglik_v1(beta, y, X, task_id),
  v2 = loglik_v2(beta, y, X, task_id),
  v3 = loglik_v3(beta, y, X, task_id))
```

Now the stopwatch:

```{r}
data.frame(
    version = c("v1 pure loop", "v2 hybrid", "v3 grouped"),
    ms = c(time_ms(loglik_v1(beta, y, X, task_id)),
           time_ms(loglik_v2(beta, y, X, task_id)),
           time_ms(loglik_v3(beta, y, X, task_id)))
)
```

The numbers vary by machine but the *structure* does not: v3 beats v1 by two to three orders of magnitude. Understanding why is more valuable than the numbers. R is an interpreted language: each line of R costs interpreter overhead *per execution*, so 4,000 trips through a loop body pay that overhead 4,000 times --- and v1 adds injury by scanning the entire `task_id` vector inside every iteration (`which(task_id == i)`), making its cost grow with the *square* of the data size. Vectorized calls --- `X %*% beta`, `rowsum()`, arithmetic on whole vectors --- pay the overhead *once* and do the actual work in compiled C loops. "Vectorization" is not magic; it is moving the loop from interpreted code into compiled code.[^1]

[^1]: The v2 hybrid shows the technique's anatomy: vectorizing the matrix product (the arithmetic bulk) captures much of the win; eliminating the residual R-level loop over groups captures the rest. When full vectorization is awkward, the hybrid is often the right stopping point --- @sec-mixed-msl's per-person structure will put us exactly there, and knowing which loop is expensive (per-task, thousands) versus tolerable (per-draw, hundreds) is the judgment this measurement builds.

Multiply the payoff through @tbl-eval-count: at v1 speed, the HB sampler of @sec-hb-mnl is an overnight-or-worse job; at v3 speed, minutes. Same computer, same mathematics.

## Precompute What Doesn't Change

A likelihood function gets called at hundreds of different $\bfbeta$ values --- but the *data* never change between calls. Anything computable from data alone must be computed once, outside the optimizer, never inside the likelihood. Our code has quietly followed this rule --- `model.matrix()` and `task_id` were always built before `optim()` was called --- but the temptation to violate it arrives disguised as tidiness ("let the function take the raw data frame; it's a nicer interface"):

```{r}
loglik_tidy_trap <- function(beta, df) {
    Xd <- model.matrix(~ brand + ram + screen + price, data = df)[, -1]
    tid <- cumsum(!duplicated(df[, c("id", "task")]))
    loglik_v3(beta, df$choice, Xd, tid)
}

c(precomputed = time_ms(loglik_v3(beta, y, X, task_id)),
  rebuilds    = time_ms(loglik_tidy_trap(beta, laptops)))
```

The convenient interface rebuilds a 12,000-row model matrix on every one of the $\sim 10^6$ calls in @tbl-eval-count. The design discipline: *functions called inside optimization loops take prepared numeric objects, full stop.* If you want the friendly interface, write a wrapper that prepares once and then loops --- exactly the division of labor in `build_choice_data()` (@sec-choice-data) followed by `loglik_mnl()`. The same principle, one level up, will matter in @sec-msl: the random draws used by a simulated likelihood are also "data" in this sense, generated once and reused across all evaluations --- there for statistical reasons as well as speed.

## Preallocate; Don't Grow

The third habit concerns memory, and its natural habitat is the MCMC chapters ahead, where samplers *store* tens of thousands of parameter draws. R objects are immutable under the hood: "appending" to a vector actually allocates a new, longer vector and copies everything across. Grow a vector inside a loop and you copy $1 + 2 + \cdots + n$ elements --- quadratic cost, invisible in the code's appearance:

```{r}
grow <- function(n) {
    out <- numeric(0)
    for (i in 1:n) out <- c(out, sqrt(i))
    out
}
prealloc <- function(n) {
    out <- numeric(n)
    for (i in 1:n) out[i] <- sqrt(i)
    out
}
c(grow_50k     = time_ms(grow(50000), times = 3),
  prealloc_50k = time_ms(prealloc(50000), times = 3))
```

Identical output, painful ratio --- and the ratio *worsens* as $n$ grows. The rule writes itself: when the number of results is known (an MCMC run of $S$ draws of $K$ parameters), allocate the full container first --- `draws <- matrix(NA_real_, nrow = S, ncol = K)` --- and fill by index. Every sampler in Parts III--IV is built this way; when you meet `matrix(NA, S, K)` at the top of the Gibbs samplers in @sec-gibbs, this paragraph is why.

A related but distinct caution: R's *copy-on-modify* semantics mean that modifying an object passed into a function, or touched by two names, can trigger a full copy. For our workloads this bites rarely --- we mostly *read* `X` inside tight loops, which is free --- but if you ever find memory ballooning in a sampler, the culprit is usually an innocent-looking modification of a large object inside a loop.

## Numerical Hygiene Is Performance Too

Speed is worthless if the fast answer is `NaN`. Two habits from earlier chapters deserve re-statement as *performance* practices, because numerical failures cost more research time than slow code ever does:

- **The log-sum-exp guard** (@sec-logsumexp) costs two vectorized operations and makes the MNL kernel safe at any $\bfbeta$ an optimizer or sampler will ever try. v3 carries it; keep it.
- **Stable scalar forms.** The binary logit log-probability $y \log P + (1-y)\log(1-P)$ is prettier as mathematics than as computation; the algebraically equal form $y v - \log(1 + e^{v})$ --- where $v = \x'\bfbeta$ is the binary model's systematic utility --- computed as `y*v - log1p(exp(v))`, avoids catastrophic cancellation for moderate $v$ (that is the form the @sec-optimization contour helper used --- promise kept). R's `log1p()` and `expm1()` exist precisely for "log of one-plus-small" and "e-to-small-minus-one" situations; reach for them whenever a probability sits near 0 or 1.

The general principle: *transform to the scale where the computer's arithmetic is happiest --- almost always the log scale --- and stay there as long as possible.* MCMC will hand us the same lesson from another angle: Metropolis acceptance ratios are computed as differences of log-posteriors, never ratios of posteriors, for exactly this reason (@sec-bayes-mnl).

## When R Isn't Enough: A Glimpse Beyond

Sometimes the hot loop cannot be vectorized --- its iterations depend on each other (the Gibbs samplers of Part III update sequentially by construction), or the group-wise logic outgrows `rowsum()`'s vocabulary. The escape hatch is to write the loop in C++ and call it from R via **Rcpp**. The flavor, for the MNL kernel (display only --- compiling C++ requires the Rtools toolchain on Windows):

```{r}
#| eval: false
Rcpp::cppFunction('
double loglik_mnl_cpp(NumericVector beta, IntegerVector y,
                      NumericMatrix X, IntegerVector task_id) {
    // one pass over rows: accumulate per-task max, log-sum, chosen V
    // ... a dozen lines of straightforward C++ ...
}')
```

Rcpp typically buys another factor of 2--20 over well-vectorized R --- valuable, not transformative, precisely *because* good vectorized R already spends most of its time in compiled code. The production-grade example in our domain is the `mixl` package [@molloy2021], which compiles each model's likelihood to C++ with OpenMP parallelism and estimates mixed logits on huge datasets; its design --- specify the model at a high level, generate compiled code underneath --- shows where the craft goes at industrial scale. For this book, we stay in plain R: our vectorized kernels keep every chapter's computation in the minutes-not-hours regime, and transparency is the whole point of the exercise. But know the door exists, and know that everything on this chapter's checklist --- measure, vectorize, precompute, preallocate --- applies unchanged on the other side of it.

## Key Learnings

- The likelihood kernel is the bottom of a multiplication stack (@tbl-eval-count): optimizers call it hundreds of times, MSL multiplies by draws, MCMC by tens of thousands of iterations. Kernel speed is research feasibility.
- Measure before optimizing --- `system.time()` medians for comparisons, `profvis` for anatomy. Intuition lies.
- Vectorization moves loops from interpreted R into compiled code: the grouped MNL kernel (matrix product + `rowsum()` + broadcast) beats the per-task loop by orders of magnitude, and the hybrid version shows which part of the win comes from where.
- Precompute the invariant: model matrices, index vectors --- and later, MSL draws --- are built once outside the loop. Functions inside optimization loops take prepared numeric objects.
- Preallocate known-size results (`matrix(NA, S, K)` atop every sampler to come); growing objects in loops is quadratic copying.
- Numerical hygiene is performance: log-sum-exp guards, `log1p()`/`expm1()`, and staying on the log scale prevent the failures that cost real time. Rcpp is the escape hatch when an unvectorizable loop truly binds --- `mixl` shows the industrial version --- but transparent, vectorized R carries this entire book comfortably.
