::: {.content-hidden}
{{< include ../latex-header.tex >}}
:::

# Relaxing IIA: Nested Logit and GEV Models {#sec-iia}

```{r}
#| label: setup
#| message: false
#| echo: false
library(ggplot2)
theme_set(theme_minimal())
```

The MNL bought its closed form with the IIA property (@eq-iia), and @sec-mnl showed the bill can be absurd: paint a bus red and steal car riders. This chapter makes the first --- and gentlest --- repair. The **nested logit** lets alternatives share unobserved appeal *within researcher-specified groups*, breaking IIA across groups while preserving a closed-form probability, a tractable likelihood, and everything we know about estimating such things. It is the smallest possible step beyond the MNL, which makes it the ideal place to learn how model generalizations propagate through the estimation machinery: a new parameter, a new likelihood, the same workflow.

The chapter also marks a small narrative shift in the running example, explained below, and introduces a technique we will use for the rest of the book: simulating data *directly from choice probabilities* when utility-plus-error simulation is awkward.

## When Proportional Substitution Fails

Recall the diagnosis from @sec-mnl: the MNL's independent errors give it no way to know that two alternatives are similar in unobserved ways, so improvement in one alternative draws share from all others *in proportion to their shares* --- never disproportionately from its close substitutes.

Put that in laptop terms. Suppose unobserved utility has a large operating-system component: some shoppers, on some occasions, are simply in a macOS mood or locked into Windows workflows we don't measure. Then a Dell and an Acer (both Windows) share unobserved appeal that neither shares with an Apple. If the Dell cuts its price, common sense says it should poach *mostly Windows shoppers* --- Acer's customers --- and pry away few committed Mac buyers. The MNL is structurally incapable of that asymmetry; the nested logit is built for it.

## The Nested Logit Model

Partition the alternatives in a choice situation into $M$ **nests** $B_1, \ldots, B_M$. Our study: $B_{\text{win}} = \{\text{Acer}, \text{Dell}\}$, $B_{\text{mac}} = \{\text{Apple}\}$ (a nest may be a singleton). Utility keeps its form $U_{ntj} = V_{ntj} + \varepsilon_{ntj}$, but the joint distribution of the errors is generalized so that errors *within* a nest are correlated while errors *across* nests remain independent. The degree of within-nest correlation is governed by the **dissimilarity parameter** $\lambda_m \in (0, 1]$: the within-nest error correlation is approximately $1 - \lambda_m^2$ --- a standard approximation we take from the literature rather than derive, since the exact correlation has no simple closed form [@train2009, Section 4.2] --- so

- $\lambda_m = 1$: no correlation --- the nest is cosmetic, and the model collapses to the MNL;
- $\lambda_m \to 0$: perfect correlation --- alternatives in the nest are near-clones on unobservables, and competition within the nest becomes ferocious while the nest as a whole competes with others as a unit.

Values outside $(0,1]$ break consistency with random utility maximization,[^1] a fact with practical bite: an estimated $\hat\lambda$ above 1 is a specification alarm (the nesting is wrong), and our estimation code below will constrain the search to the valid interval. For this chapter we give both nests a *common* $\lambda$ --- the natural choice when one nest is a singleton, whose own $\lambda$ would be unidentified.

[^1]: @train2009 Section 4.2 discusses the bounds and the (limited) sense in which $\lambda$ slightly above 1 can be tolerated. The formal foundation --- that the nested logit is the GEV member generated by a particular function $G$ --- appears at the chapter's end.

The choice probability [derived in @train2009, Section 4.2] has a structure that teaches the model's behavioral content. Define the **inclusive value** of nest $m$:

$$
I_{ntm} = \log \sum_{j' \in B_m} e^{V_{ntj'} / \lambda}
$$ {#eq-inclusive-value}

Then

$$
P_{ntj} = \underbrace{\frac{e^{\lambda I_{ntm(j)}}}{\sum_{m'=1}^{M} e^{\lambda I_{ntm'}}}}_{\Pr(\text{nest } m(j))} \times \underbrace{\frac{e^{V_{ntj}/\lambda}}{\sum_{j' \in B_{m(j)}} e^{V_{ntj'}/\lambda}}}_{\Pr(j \mid \text{nest } m(j))}
$$ {#eq-nested-prob}

Read it as a two-stage decision --- pick a nest, then pick within the nest --- with each stage an MNL. The inclusive value is the bridge between stages: $I_{ntm}$ is the log-sum of nest $m$, the same "value of a choice set" object that the Gumbel max-stability property produced in @sec-mnl, now measuring how attractive the nest is *as a whole* when the upper stage compares nests.[^2] The two-stage reading is a mathematical decomposition, not a claim about sequential cognition; the model is a simultaneous utility maximization with correlated errors that happens to factor this way.

[^2]: The log-sum's third appearance is scheduled for @sec-postestimation, where (divided by the price coefficient) it becomes consumer surplus. One object, three jobs: max-stability byproduct, nest attractiveness, welfare measure.

Verify the two limiting cases against @eq-nested-prob by inspection: at $\lambda = 1$, the $I$'s become ordinary log-sums and the product collapses to @eq-mnl-prob; as $\lambda \to 0$, the within-nest stage approaches "pick the nest's best alternative with certainty" while nests compete by their best members. And confirm the IIA repair by forming the odds for $j \in B_a$ and $k \in B_b$ with $a \neq b$. The across-nest denominator is common to both probabilities and cancels, and using the identity $\sum_{j' \in B_m} e^{V_{ntj'}/\lambda} = e^{I_{ntm}}$ for each within-nest denominator,
$$\frac{P_{ntj}}{P_{ntk}} = \frac{e^{\lambda I_{nta}} \; e^{V_{ntj}/\lambda - I_{nta}}}{e^{\lambda I_{ntb}} \; e^{V_{ntk}/\lambda - I_{ntb}}} = e^{(V_{ntj} - V_{ntk})/\lambda} \; e^{(\lambda - 1)(I_{nta} - I_{ntb})}.$$
At $\lambda = 1$ the second factor is 1 and we are back to @eq-iia. For $\lambda < 1$, the odds depend --- through $I_{nta}$ and $I_{ntb}$ --- on every alternative in the *two nests involved*: improve the Acer and the Dell/Apple odds shift, because $I_{\text{win}}$ rises. Note the precision of the repair: alternatives in any *third* nest still cancel exactly, so IIA fails at the boundary between the two nests being compared, survives within nests, and survives with respect to everything else. Choosing the boundary is the researcher's model specification burden --- and the model offers no protection against choosing it wrongly, a limitation the mixed logit (@sec-mixed-msl) and probit (@sec-mnp) will lift at the cost of the closed form.

## A Labeled Variant of the Study

Nesting is a statement about *which alternatives* share unobserved appeal, so it attaches most naturally to designs where the alternatives have stable identities. For this chapter we switch the running example to a **brand-labeled design**: every task shows exactly one Acer, one Dell, and one Apple --- alternative 1 *is* the Acer, and so on --- with memory, screen, and price still randomized independently. This is a common real-world CBC variant (@sec-data introduced the labeled/unlabeled distinction; here it earns its keep), and it changes the bookkeeping in one satisfying way: the brand dummies of @sec-mnl are now, literally, **alternative-specific constants** --- the ASCs that @sec-data-intercept said labeled designs require. Same columns, same true values, new interpretation. The utility specification and true parameters carry over from @eq-beta-true, joined by the new structural parameter:

$$
\lambda^\ast = 0.5
$$

--- substantial within-nest correlation (about $1 - 0.5^2 = 0.75$), enough to make the substitution patterns visibly non-MNL.

## Simulating Nested Logit Data {#sec-sim-nested}

Here we hit an instructive wall. Our simulators so far follow the utility recipe: draw errors, add, argmax (@sec-simdata, @sec-sim-mnl). For the nested logit, the required error draws --- marginally Gumbel, dependent within nests in the precise GEV pattern --- have no simple constructive recipe like `rlogis()` or the Cholesky transform.[^3] But we don't need one: **any model whose choice probabilities we can compute can be simulated directly from those probabilities.** Compute $P_{ntj}$ for the task via @eq-nested-prob, then draw the choice from the corresponding categorical distribution --- in R, `sample()` with `prob =`. The utility framework earns its keep in *deriving* the probabilities; once derived, they are the model.

[^3]: Constructions exist [@train2009, Section 4.2 notes the possibilities], but they are genuinely awkward, and the probability route is both exact and simpler. The probability route also has a conceptual payoff: it separates "the model" (a distribution over choices given attributes) from "the story that motivated it" (utilities and errors).

```{r}
sim_nested_data <- function(N, T, beta, lambda) {
    J <- 3
    n_rows <- N * T * J
    df <- data.frame(
        id    = rep(seq_len(N), each = T * J),
        task  = rep(rep(seq_len(T), each = J), times = N),
        alt   = rep(c("Acer", "Dell", "Apple"), times = N * T),
        ram   = factor(sample(c("8","16","32"), n_rows, replace = TRUE),
                       levels = c("8","16","32")),
        screen = factor(sample(c("13","15"), n_rows, replace = TRUE),
                        levels = c("13","15")),
        price = round(runif(n_rows, min = 0.8, max = 2.4), 2)
    )
    df$alt  <- factor(df$alt, levels = c("Acer", "Dell", "Apple"))
    df$nest <- ifelse(df$alt == "Apple", "mac", "win")

    X <- model.matrix(~ alt + ram + screen + price, data = df)[, -1]
    V <- as.vector(X %*% beta)
    task_id <- cumsum(!duplicated(df[, c("id", "task")]))

    ## nested logit probabilities, task by task (eq-nested-prob)
    P <- numeric(n_rows)
    for (i in unique(task_id)) {
        rows <- which(task_id == i)
        v    <- V[rows]; nest <- df$nest[rows]
        Iv   <- tapply(v / lambda, nest, function(z) log(sum(exp(z))))
        Pnest   <- exp(lambda * Iv) / sum(exp(lambda * Iv))
        Pwithin <- exp(v / lambda) / ave(exp(v / lambda), nest, FUN = sum)
        P[rows] <- Pnest[nest] * Pwithin
    }
    ## choices: one categorical draw per task
    df$choice <- 0L
    for (i in unique(task_id)) {
        rows <- which(task_id == i)
        df$choice[rows[sample.int(J, 1, prob = P[rows])]] <- 1L
    }
    df
}

set.seed(103)
beta_true   <- c(dell = 0.5, apple = 1.0, ram16 = 0.6, ram32 = 0.9,
                 screen15 = 0.3, price = -1.2)
lambda_true <- 0.5
nested <- sim_nested_data(N = 500, T = 8, beta = beta_true,
                          lambda = lambda_true)
```

(Yes, the simulator loops over tasks --- and by @sec-performance's own doctrine that is fine: it runs *once*, not inside an optimizer. Measure before optimizing includes knowing when not to bother.)

### Seeing the repaired substitution

Before estimating, let's confirm the model does what it was hired to do. Take a representative task --- one laptop per brand, identical specs, \$1,500 each --- cut the Dell's price by \$300, and ask where its gained share comes from, under the nested model ($\lambda = 0.5$) versus the MNL ($\lambda = 1$):

```{r}
nested_probs <- function(V, nest, lambda) {
    Iv      <- tapply(V / lambda, nest, function(z) log(sum(exp(z))))
    Pnest   <- exp(lambda * Iv) / sum(exp(lambda * Iv))
    Pwithin <- exp(V / lambda) / ave(exp(V / lambda), nest, FUN = sum)
    as.vector(Pnest[nest] * Pwithin)
}

X_task <- rbind(acer  = c(0, 0, 1, 0, 1, 1.5),
                dell  = c(1, 0, 1, 0, 1, 1.5),
                apple = c(0, 1, 1, 0, 1, 1.5))
nest   <- c("win", "win", "mac")
V0 <- as.vector(X_task %*% beta_true)
V1 <- V0 + c(0, beta_true["price"] * -0.3, 0)   # Dell $300 cheaper: price falls by 0.3 ($000s)

shares <- rbind(
    nested_before = nested_probs(V0, nest, 0.5),
    nested_after  = nested_probs(V1, nest, 0.5),
    mnl_before    = nested_probs(V0, nest, 1),
    mnl_after     = nested_probs(V1, nest, 1)
)
colnames(shares) <- rownames(X_task)
round(shares, 3)
```

Do the accounting. Under the MNL, the Dell's gain comes from Acer and Apple in proportion to their prior shares --- proportional substitution, as @sec-mnl proved it must. Under the nested model, Acer surrenders share at a far higher *rate* relative to its starting point than Apple does: the price cut plays out mostly as within-nest warfare among the Windows machines, while the Mac buyer barely glances up. That asymmetry --- impossible in the MNL --- is the entire point of the $\lambda$ parameter.

## The Nested Logit Log-Likelihood

Estimation needs $\log P_{ntj}$ for the chosen alternatives. One identity does all the work: by the definition @eq-inclusive-value, the within-nest denominator is $\sum_{j' \in B_{m(j)}} e^{V_{ntj'}/\lambda} = e^{I_{ntm(j)}}$. So the log of @eq-nested-prob's within-nest factor is $V_{ntj}/\lambda - I_{ntm(j)}$, the log of its nest-choice factor is $\lambda I_{ntm(j)} - \log \sum_{m'} e^{\lambda I_{ntm'}}$, and (writing $m(j)$ for the chosen row's nest) their sum is:

$$
\log P_{ntj} = \underbrace{\left[ \frac{V_{ntj}}{\lambda} - I_{ntm(j)} \right]}_{\text{within-nest}} \; + \; \underbrace{\left[ \lambda I_{ntm(j)} - \log \textstyle\sum_{m'} e^{\lambda I_{ntm'}} \right]}_{\text{across-nest}}
$$ {#eq-nested-loglik-term}

Both bracketed terms are "utility minus log-sum" shapes --- the MNL pattern of @eq-mnl-loglik-compact at two levels. So the vectorized implementation is the @sec-mnl-mle machinery applied twice: once with rows grouped by *(task, nest)* to build inclusive values, once with nests grouped by *task*. The parameter vector stacks the two: $\bftheta = (\bfbeta, \lambda)$ --- `theta` in the code, unpacked by `beta <- theta[1:K]` and `lambda <- theta[K + 1]`; each level gets its own max-shift for numerical stability (the log-sum-exp guard of @sec-logsumexp, applied twice):

```{r}
loglik_nested <- function(theta, y, X, task_id, nest_id) {
    K      <- ncol(X)
    beta   <- theta[1:K]
    lambda <- theta[K + 1]
    Vs <- as.vector(X %*% beta) / lambda

    ## level 1: inclusive value per (task, nest) group
    g      <- as.integer(interaction(task_id, nest_id, drop = TRUE))
    m1_row <- ave(Vs, g, FUN = max)
    m1     <- as.vector(tapply(m1_row, g, `[`, 1))
    I_g    <- log(as.vector(rowsum(exp(Vs - m1_row), g))) + m1
    task_of_g <- as.vector(tapply(task_id, g, `[`, 1))

    ## level 2: log-sum of lambda * I across nests within task
    lI   <- lambda * I_g
    m2_g <- ave(lI, task_of_g, FUN = max)
    m2   <- as.vector(tapply(m2_g, task_of_g, `[`, 1))
    lse2 <- log(as.vector(rowsum(exp(lI - m2_g), task_of_g))) + m2

    ## eq-nested-loglik-term, chosen rows only
    logP <- (Vs - I_g[g]) + (lambda * I_g[g] - lse2[task_id])
    sum(logP[y == 1])
}
```

The new idiom is `interaction(task_id, nest_id, drop = TRUE)`: it mints one group label per (task, nest) pair, letting `rowsum()` and `ave()` aggregate at the intermediate level, after which `I_g[g]` and `lse2[task_id]` broadcast group-level quantities back down --- the same label-indexing trick as always, used at two altitudes. Check it the honest way, against a brute-force task-by-task computation of @eq-nested-prob at the true parameters:

```{r}
X <- model.matrix(~ alt + ram + screen + price, data = nested)[, -1]
y <- nested$choice
task_id <- cumsum(!duplicated(nested[, c("id", "task")]))

ll_fast <- loglik_nested(c(beta_true, lambda_true), y, X, task_id, nested$nest)

V <- as.vector(X %*% beta_true); ll_bf <- 0
for (i in unique(task_id)) {
    rows <- which(task_id == i)
    P <- nested_probs(V[rows], nested$nest[rows], lambda_true)
    ll_bf <- ll_bf + log(P[y[rows] == 1])
}
c(vectorized = ll_fast, brute_force = ll_bf)
```

Identical. Note in passing what we just did: *the simulator and the likelihood were implemented independently* (probability formula task-by-task versus grouped log-sum algebra), so their agreement is a genuine cross-check, not a tautology --- the same reason @sec-mnl-mle validated against `mlogit` rather than against itself.

## Estimation

One wrinkle separates this fit from @sec-mnl-mle: $\lambda$ must stay in $(0, 1]$, and BFGS knows nothing of bounds. Three standard options: reparameterize (estimate $\alpha$ with $\lambda = 1/(1+e^{-\alpha})$, unconstrained), penalize, or use a bounded optimizer. We take the third --- `optim(method = "L-BFGS-B")` accepts box constraints directly, and meeting it here matters because bounded parameters recur (class shares in @sec-latentclass, variance parameters in @sec-mixed-msl, where we will choose reparameterization instead and compare notes):

```{r}
fit <- optim(par     = c(rep(0, 6), 0.8),
             fn      = loglik_nested,
             y = y, X = X, task_id = task_id, nest_id = nested$nest,
             method  = "L-BFGS-B",
             lower   = c(rep(-Inf, 6), 0.05),
             upper   = c(rep( Inf, 6), 1),
             control = list(fnscale = -1),
             hessian = TRUE)
se <- sqrt(diag(solve(-fit$hessian)))

results <- data.frame(truth    = c(beta_true, lambda = lambda_true),
                      estimate = fit$par, se = se, z = fit$par / se)
rownames(results) <- c(colnames(X), "lambda")
round(results, 3)
```

Recovery across the board, including the structural parameter: $\hat\lambda \approx `r round(fit$par[7], 2)`$ with a tight standard error, decisively inside $(0,1)$ and decisively away from the MNL's $\lambda = 1$. Make that last comparison formal with the likelihood-ratio machinery of @sec-mnl-mle --- the MNL is this model with one parameter pinned, so:

```{r}
loglik_mnl <- function(beta, y, X, task_id) {      # from @sec-mnl-mle
    V     <- as.vector(X %*% beta)
    Vmax  <- ave(V, task_id, FUN = max)
    eV    <- exp(V - Vmax)
    sum((V - Vmax - log(rowsum(eV, task_id)[task_id]))[y == 1])
}
fit_mnl <- optim(rep(0, 6), loglik_mnl, y = y, X = X, task_id = task_id,
                 method = "BFGS", control = list(fnscale = -1))
lr <- 2 * (fit$value - fit_mnl$value)
c(LR = round(lr, 1), p = signif(pchisq(lr, df = 1, lower.tail = FALSE), 3))
```

The data, simulated with real nest correlation, reject the MNL emphatically.[^4] Estimating a wrong-by-construction MNL on these data is itself instructive --- compare `fit_mnl$par` to the truth and you'll find the *attribute* coefficients survive respectably while the substitution predictions (the model's whole purpose in a competitive analysis) would be badly off: misspecification often hides in the quantities you don't tabulate.

[^4]: A technical footnote for the careful: the null $\lambda = 1$ sits on the *boundary* of the parameter space, which makes the usual $\chi^2_1$ reference conservative in general. At an LR statistic this size, no correction changes the verdict.

## Validation Against `mlogit`

`mlogit` estimates nested logits via its `nests` argument; `un.nest.el = TRUE` imposes our common-$\lambda$ assumption. With labeled alternatives, `mlogit` supplies the ASCs itself (its `(Intercept):Dell` and `(Intercept):Apple` are our `dell` and `apple` dummies --- same reference brand, Acer):

```{r}
#| message: false
library(dfidx)
library(mlogit)

md <- nested
md$chid   <- task_id
md$choice <- as.logical(md$choice)
md <- dfidx(md, idx = list(c("chid", "id"), "alt"))

fit_ml <- mlogit(choice ~ ram + screen + price, data = md,
                 nests = list(win = c("Acer", "Dell"), mac = "Apple"),
                 un.nest.el = TRUE)

round(rbind(ours = fit$par, mlogit = coef(fit_ml)[c(1:2, 3:6, 7)]), 4)
c(ours = fit$value, mlogit = as.numeric(logLik(fit_ml)))
```

Coefficient-for-coefficient agreement (their `iv` is our $\lambda$), identical log-likelihoods. Two implementations, one summit, again.

## The GEV Family {#sec-gev-family}

The nested logit is one member of a family McFadden constructed with a single generating device [@mcfadden2001; @train2009, Ch. 4]. Take any function $G(e^{V_1}, \ldots, e^{V_J})$ satisfying a short list of properties (non-negativity, homogeneity, sign-alternating derivatives); then

$$
P_j = \frac{e^{V_j} \, G_j(e^{V_1}, \ldots, e^{V_J})}{G(e^{V_1}, \ldots, e^{V_J})}
$$ {#eq-gev-prob}

(with $G_j = \partial G / \partial e^{V_j}$) is the choice probability of a well-defined random utility model with **generalized extreme value** errors --- closed-form, RUM-consistent, guaranteed. The family is a probability factory: $G = \sum_j e^{V_j}$ yields the MNL; $G = \sum_m \big( \sum_{j \in B_m} e^{V_j/\lambda_m} \big)^{\lambda_m}$ yields the nested logit (differentiate it and @eq-nested-prob falls out); overlapping-nest choices of $G$ yield the *cross-nested* and *paired combinatorial* logits, where an alternative belongs partially to several nests --- a Dell desktop-replacement laptop that competes in both the "Windows" nest and a "large-screen" nest.

We won't estimate these siblings --- the workflow is exactly the one this chapter just executed: derive $P$ from $G$, take logs, write the two-(or three-)level grouped likelihood, constrain the $\lambda$'s, climb --- and one worked example plus a recipe beats three worked examples. The strategic lesson to carry forward instead: within the GEV family, *every* substitution pattern must be specified in advance through the structure of $G$. The researcher declares who competes with whom; the data only measure how strongly. When you do not know the competitive structure --- the usual situation --- you want models where flexible substitution *emerges* from estimated quantities instead: random coefficients (@sec-mixed-msl) or free error covariance (@sec-mnp). Both cost the closed form. The next four chapters pay that cost and build the simulation machinery that covers it.

## Key Learnings

- The nested logit groups alternatives into nests with correlated unobservables; the dissimilarity parameter $\lambda \in (0,1]$ indexes the correlation ($\lambda = 1$ recovers the MNL), and estimates outside the interval indict the nesting structure itself.
- The probability (@eq-nested-prob) factors into two MNL stages linked by the inclusive value --- the log-sum of @sec-mnl in its second starring role --- and IIA now fails across nests while surviving within them, exactly as designed: our \$300 Dell price cut savaged Acer and barely touched Apple.
- New simulation technique: when errors have no convenient constructive form, simulate *from the choice probabilities directly* (categorical draws). Computable probabilities are all a simulator needs.
- The log-likelihood is the MNL's "utility minus log-sum" pattern applied at two levels; `interaction()` group labels plus `rowsum()`/broadcast implement it vectorized, with max-shifts at both levels. Simulator and likelihood, implemented independently, agreed at the truth --- and `mlogit` agreed with both.
- Bounded parameters met bounded optimization: `L-BFGS-B` with box constraints. The LR test rejected $\lambda = 1$ decisively (mind the boundary caveat), and recovery included the structural parameter.
- The GEV generating function (@eq-gev-prob) mass-produces closed-form RUM models --- but all of them require the researcher to pre-specify who competes with whom. Flexible, data-driven substitution requires integrals without closed forms: enter simulation-assisted estimation.
