::: {.content-hidden}
{{< include ../latex-header.tex >}}
:::

# The Latent Class MNL Model {#sec-latentclass}

```{r}
#| label: setup
#| message: false
#| echo: false
library(ggplot2)
theme_set(theme_minimal())
```

Every model so far has made an assumption so ingrained you may not have registered it as one: *everybody has the same $\bfbeta$.* One price coefficient for the student and the executive, one Apple premium for the designer and the accountant. For the questions that make choice modeling valuable in practice --- who is price-sensitive, which segments value which features, how heterogeneous is demand --- that assumption is not a simplification; it assumes away the subject.

This chapter introduces preference **heterogeneity** in its most tractable form: a finite number of unobserved *classes*, each with its own coefficient vector. The latent class MNL matters for three reasons. Substantively, it is the model behind data-driven market segmentation. Technically, it forces us to confront the panel structure of our data --- eight choices per respondent are *not* eight independent observations once people differ --- and the likelihood repair we make here is *exactly* the one the mixed logit needs in @sec-mixed-msl. And algorithmically, it introduces both the EM algorithm and, through the E-step's posterior class probabilities, this book's first genuine application of Bayes' rule --- a quiet rehearsal for Part III.

## The Model

Suppose the population consists of $C$ unobserved classes. A decision-maker in class $c$ chooses according to a standard MNL with class-specific coefficients $\bfbeta_c$; the probability that a randomly sampled person belongs to class $c$ is the **class share** $\pi_c$, with $\sum_c \pi_c = 1$. The model's parameters stack the shares and all the class coefficient vectors:

$$
\bftheta = (\pi_1, \ldots, \pi_{C-1}, \; \bfbeta_1, \ldots, \bfbeta_C)
$$

Membership is *latent* --- never observed, neither in real data nor (we pretend) in ours. What does the model imply for what we *do* observe, a person's whole sequence of choices?

## The Likelihood: Where the Panel Structure Bites

Here is the reasoning, slowly, because a subtle error lurks and nearly everyone makes it once. Condition on person $n$ belonging to class $c$: their $T$ choices are then independent MNL draws (the errors $\varepsilon_{ntj}$ are still iid), so the probability of their observed *sequence* $\y_n = (y_{n1}, \ldots, y_{nT})$ is a product over tasks:

$$
\Pr(\y_n \mid \text{class } c) = \prod_{t=1}^{T} P_{nt}(\bfbeta_c)
$$

where $P_{nt}(\bfbeta_c)$ is the MNL probability (@eq-mnl-prob) of the alternative person $n$ chose in task $t$, evaluated at $\bfbeta_c$. Class membership is unknown, so the unconditional probability of the sequence mixes over classes:

$$
\Pr(\y_n \mid \bftheta) = \sum_{c=1}^{C} \pi_c \prod_{t=1}^{T} P_{nt}(\bfbeta_c)
$$ {#eq-lc-person}

and the sample log-likelihood sums over *people*:

$$
\ell(\bftheta) = \sum_{n=1}^{N} \log \left[ \sum_{c=1}^{C} \pi_c \prod_{t=1}^{T} P_{nt}(\bfbeta_c) \right]
$$ {#eq-lc-loglik}

**The mixture sits outside the product over tasks.** A person is in one class for all eight of their tasks --- their choices are linked by shared membership --- so we mix *sequence* probabilities, not task probabilities. The tempting wrong version, $\sum_{n,t} \log \sum_c \pi_c P_{nt}(\bfbeta_c)$, mixes inside: it describes a different (and rarely intended) world where every person re-rolls their class before every task, and it discards precisely the information that identifies the classes --- *consistency of behavior within person*. One person choosing cheap laptops eight times in a row is strong evidence a price-sensitive class exists; eight scattered cheap choices by different people are not. Getting this hierarchy right is the single transferable skill of this chapter: the mixed logit's likelihood (@sec-mixed-msl) has identical structure with an integral in place of the sum, and the HB models of Part III have it too.

Note what @eq-lc-person also implies: the latent class model breaks IIA at the population level. Each class is an IIA-obeying MNL, but the *mixture* of them is not --- classes with different tastes generate exactly the correlated-across-alternatives behavior that @sec-iia engineered through nests, now emerging from heterogeneity instead of being specified. This is the first glimpse of a theme that culminates in the mixed logit: heterogeneity is a substitution-pattern engine.

## Simulating Latent Class Data

The story writes the simulator: assign each person a class with probabilities $(\pi_1, \ldots, \pi_C)$, then run the standard MNL simulator of @sec-sim-mnl with that person's class coefficients. We invent two segments with a familiar marketing flavor --- a majority of *premium seekers* (strong brand and memory preferences, mild price sensitivity) and a minority of *value shoppers* (weak feature preferences, fierce price sensitivity):

```{r}
#| echo: false
sim_laptop_design <- function(N, T, J) {
    n_rows <- N * T * J
    data.frame(
        id    = rep(seq_len(N), each = T * J),
        task  = rep(rep(seq_len(T), each = J), times = N),
        alt   = rep(seq_len(J), times = N * T),
        brand = factor(sample(c("Acer","Dell","Apple"), n_rows, replace = TRUE),
                       levels = c("Acer","Dell","Apple")),
        ram   = factor(sample(c("8","16","32"), n_rows, replace = TRUE),
                       levels = c("8","16","32")),
        screen = factor(sample(c("13","15"), n_rows, replace = TRUE),
                        levels = c("13","15")),
        price = round(runif(n_rows, min = 0.8, max = 2.4), 2)
    )
}
```

```{r}
sim_lc_data <- function(N, T, J, beta_list, shares) {
    df <- sim_laptop_design(N, T, J)          # the @sec-data design generator
    X  <- model.matrix(~ brand + ram + screen + price, data = df)[, -1]
    task_id <- cumsum(!duplicated(df[, c("id", "task")]))

    cls <- sample(seq_along(shares), N, replace = TRUE, prob = shares)
    beta_row <- do.call(rbind, beta_list)[cls[df$id], ]   # each row: its person's class beta
    V   <- rowSums(X * beta_row)
    eps <- -log(-log(runif(nrow(df))))
    U   <- V + eps
    df$choice <- as.integer(ave(U, task_id, FUN = function(u) u == max(u)))
    attr(df, "class_assign") <- cls                        # the answer key
    df
}

set.seed(110)
beta_premium <- c(dell = 0.6, apple = 1.5, ram16 = 0.8, ram32 = 1.4,
                  screen15 = 0.5, price = -0.8)
beta_value   <- c(dell = 0.3, apple = 0.2, ram16 = 0.3, ram32 = 0.4,
                  screen15 = 0.1, price = -2.0)
shares_true  <- c(premium = 0.6, value = 0.4)

lc <- sim_lc_data(N = 500, T = 8, J = 3,
                  beta_list = list(beta_premium, beta_value),
                  shares = shares_true)
```

Two implementation notes. The line `beta_row <- do.call(rbind, beta_list)[cls[df$id], ]` builds a matrix with *each row's own class coefficients* by double indexing --- person of the row, class of the person --- so `rowSums(X * beta_row)` computes every utility with its correct $\bfbeta_c$ in one vectorized stroke; this row-specific-coefficients idiom returns in @sec-mixed-msl where every *person* has their own $\bfbeta_n$. And the true class assignments ride along as an attribute: unavailable to any estimator, but waiting to grade our posterior classifications later.

A directional sanity check (@sec-sanity-checks): true premium seekers should buy more Apples and pricier machines than true value shoppers ---

```{r}
cls   <- attr(lc, "class_assign")
chose <- lc[lc$choice == 1, ]
data.frame(
    class       = c("premium", "value"),
    share_apple = round(tapply(chose$brand == "Apple", cls[chose$id], mean), 2),
    avg_price   = round(tapply(chose$price, cls[chose$id], mean), 2)
)
```

--- and emphatically they do.

## Route 1: Direct Maximization

Nothing stops us from feeding @eq-lc-loglik to `optim()`, and it is the honest baseline. Two mechanical matters first.

*Constrained shares.* The $\pi_c$ live on the simplex (positive, sum to one). Rather than constrain the optimizer (the @sec-iia approach), this time we **reparameterize** --- the other standard trick: optimize unconstrained scores $a_2, \ldots, a_C$ (with $a_1 = 0$ fixed) and map them through a softmax, $\pi_c = e^{a_c} / \sum_l e^{a_l}$. Any real-valued scores yield valid shares, so the constraint simply disappears from the optimizer's world.[^1]

[^1]: Both devices --- bounded optimizers and reparameterization --- appear in this book so you own both. Rules of thumb: reparameterize simplexes and "must be positive" scales (log-parameterize); use `L-BFGS-B` for simple intervals like the nested logit's $\lambda$. Remember that after reparameterization, standard errors for the *transformed* quantities need the delta method (@sec-postestimation).

*Per-person likelihoods.* @eq-lc-loglik needs, for each person and class, the log of their sequence probability $\sum_t \log P_{nt}(\bfbeta_c)$ --- our stacked machinery delivers this with one more `rowsum()`: per-task chosen log-probabilities (the @sec-mnl-mle kernel), then summed within person. And mixing the classes on the log scale is a log-sum-exp with weights --- max-shifted, per @sec-logsumexp, because sequence probabilities of $T = 8$ choices are small enough to underflow when exponentiated carelessly:

```{r}
persll_mnl <- function(beta, y, X, task_id, id_of_task) {
    V    <- as.vector(X %*% beta)
    Vmax <- ave(V, task_id, FUN = max)
    eV   <- exp(V - Vmax)
    logP <- (V - Vmax - log(rowsum(eV, task_id)[task_id]))[y == 1]  # per task
    as.vector(rowsum(logP, id_of_task))                             # per person
}

loglik_lc <- function(theta, y, X, task_id, id_of_task, n_id, C) {
    K     <- ncol(X)
    a     <- c(0, theta[1:(C - 1)])
    pi_c  <- exp(a) / sum(exp(a))
    betas <- matrix(theta[C:(C - 1 + C*K)], nrow = C, byrow = TRUE)

    ll_nc <- sapply(1:C, function(c)
        persll_mnl(betas[c, ], y, X, task_id, id_of_task))   # N x C matrix
    M <- apply(ll_nc, 1, max)                                 # rowwise max-shift
    sum(M + log(exp(ll_nc - M) %*% pi_c))                     # eq-lc-loglik
}
```

Set up the data objects and maximize, starting --- deliberately --- from a *perturbed* neighborhood of the truth; the reason for that charity becomes the next section's subject:

```{r}
X <- model.matrix(~ brand + ram + screen + price, data = lc)[, -1]
y <- lc$choice
task_id    <- cumsum(!duplicated(lc[, c("id", "task")]))
id_of_task <- lc$id[!duplicated(lc[, c("id", "task")])]
n_id <- max(lc$id)

set.seed(111)
start <- c(0, beta_premium + rnorm(6, 0, 0.3), beta_value + rnorm(6, 0, 0.3))
fit_direct <- optim(start, loglik_lc, y = y, X = X, task_id = task_id,
                    id_of_task = id_of_task, n_id = n_id, C = 2,
                    method = "BFGS",
                    control = list(fnscale = -1, maxit = 500))
pi_hat <- exp(c(0, fit_direct$par[1])); pi_hat <- pi_hat / sum(pi_hat)

round(rbind(truth_premium = beta_premium, est_class1 = fit_direct$par[2:7],
            truth_value   = beta_value,   est_class2 = fit_direct$par[8:13]), 2)
round(c(share_class1 = pi_hat[1], truth = 0.6), 3)
```

Recovery: both coefficient vectors and the 60/40 split, from choices alone --- the model heard two voices in the data and transcribed both. (The value class's small coefficients wobble more; only $\sim 40\%$ of respondents inform them, and diffuse preferences are intrinsically harder to pin down. The standard errors, if we computed them, would say so.)

## The Terrain: Local Maxima and Label Switching

The charity of a near-truth start hid something important. The latent class log-likelihood is **not concave** --- the global-concavity luxury of the MNL (@sec-optimization's footnote) expired the moment a sum appeared inside the log. Two pathologies follow, one benign, one not.

**Label switching** is the benign one: swap the two classes' labels (and their shares) and @eq-lc-loglik is unchanged --- the model is invariant to renaming, so *every* maximum comes in $C!$ mirror-image copies. Harmless for fit, confusing for interpretation and for comparing across runs; the standard hygiene is a reporting convention, such as ordering classes by their price coefficient.

**Genuine local maxima** are not benign. Start the optimizer badly and it can climb to a summit that is not *the* summit --- often one where two classes have collapsed into near-duplicates while a real segment goes unmodeled. Watch it happen; from a start where both classes are the same flat guess:

```{r}
fit_bad <- optim(c(0, rep(0.1, 12)), loglik_lc, y = y, X = X,
                 task_id = task_id, id_of_task = id_of_task,
                 n_id = n_id, C = 2, method = "BFGS",
                 control = list(fnscale = -1, maxit = 500))
c(good_start = fit_direct$value, bad_start = fit_bad$value)
```

Same data, same model, same algorithm, two different answers --- and only the log-likelihood comparison reveals which climb ended on the true summit. The defense is the multiple-starts discipline promised in @sec-optimization: launch from several dispersed random starts, keep the best, and *report* having done so. This is not optional garnish for mixture models; it is part of the estimator.

## Route 2: The EM Algorithm

There is a second, beautiful route to the same summit, built from a thought experiment: *estimation would be easy if we could see the class labels.* Known labels would split the sample into $C$ subsamples, each fit by an ordinary MNL, with shares estimated by counting. The labels are missing --- so the **EM algorithm** [Expectation--Maximization; see @train2009, Section 14.3, for the choice-model version] iterates between guessing them probabilistically and estimating as if the guess were data:

Before the two steps, the object that justifies them. If we *could* see labels, person $n$ in class $c$ would contribute $\log \pi_c + \sum_t \log P_{nt}(\bfbeta_c)$ to a complete-data log-likelihood. We can't --- so EM averages that contribution over our current beliefs about the labels, the weights $w_{nc}$ defined below, giving the surrogate objective
$$Q(\bftheta) = \sum_n \sum_c w_{nc} \left[ \log \pi_c + \sum_t \log P_{nt}(\bfbeta_c) \right].$$
The sum splits cleanly, and each piece hands us one update. The $\bfbeta_c$ piece is, for each class, exactly a *weighted* MNL log-likelihood --- each person's tasks entering with weight $w_{nc}$. The $\pi_c$ piece, maximized subject to $\sum_c \pi_c = 1$, gives $\pi_c = \frac{1}{N}\sum_n w_{nc}$ --- the familiar "observed shares are the MLE of category probabilities," with soft counts in place of hard ones. And maximizing $Q$ is what guarantees the log-likelihood itself cannot fall from one iteration to the next.

**E-step.** Given current parameters, compute each person's *posterior class probabilities*. This is Bayes' rule --- making its first working appearance in this book --- applied to the question "given what this person chose, which class are they in?". Two lines get us there. The probability that person $n$ is in class $c$ *and* makes the choices we saw is prior times conditional: $\Pr(\text{class } c, \y_n) = \pi_c \prod_t P_{nt}(\bfbeta_c)$. Divide by the total probability of what we saw --- the joint summed over all classes, which is exactly the person's likelihood contribution @eq-lc-person --- and Bayes' rule falls out:

$$
w_{nc} = \Pr(\text{class } c \mid \y_n) = \frac{\pi_c \prod_t P_{nt}(\bfbeta_c)}{\sum_l \pi_l \prod_t P_{nt}(\bfbeta_l)}
$$ {#eq-estep}

Prior share times likelihood of the person's sequence, normalized --- and the denominator is @eq-lc-person itself. Every ingredient is already computed by `persll_mnl()`.

**M-step.** Re-estimate each class's $\bfbeta_c$ by a *weighted* MNL --- every person participates in every class's estimation, weighted by their membership probability $w_{nc}$ --- and re-estimate shares as average memberships --- both updates read directly off $Q$. The weighted MNL is our @sec-mnl-mle kernel with one multiplication added (and its gradient keeps the errors-times-covariates form, weighted):

```{r}
loglik_mnl_w <- function(beta, y, X, task_id, w_task) {
    V    <- as.vector(X %*% beta)
    Vmax <- ave(V, task_id, FUN = max)
    eV   <- exp(V - Vmax)
    logP <- (V - Vmax - log(rowsum(eV, task_id)[task_id]))[y == 1]
    sum(w_task * logP)
}
grad_mnl_w <- function(beta, y, X, task_id, w_task) {
    V    <- as.vector(X %*% beta)
    Vmax <- ave(V, task_id, FUN = max)
    eV   <- exp(V - Vmax)
    p    <- eV / rowsum(eV, task_id)[task_id]
    as.vector(crossprod(X, w_task[task_id] * (y - p)))
}
```

The full algorithm alternates the two steps until the log-likelihood stops improving. EM's celebrated guarantee --- the $Q$-maximization argument above: *each iteration cannot decrease the log-likelihood*, so it climbs steadily --- no step sizes, no line searches, no Hessians --- though toward whichever summit its start commits it to (EM shares the local-maxima problem; multiple starts still apply).

One junction in the implementation deserves a warning flag before you read it, because it is this chapter's classic bug site: the E-step's weights `w` are *per person* ($N \times C$), but the weighted likelihood consumes weights *per task*. The mapping runs through `id_of_task` --- the person who owns each task --- so the weight vector handed to the M-step is `w[id_of_task, c]`, one entry per task, each task inheriting its owner's membership probability. Index per-person quantities by the wrong level and nothing errors; the classes simply come out subtly, silently wrong.

```{r}
em_lc <- function(y, X, task_id, id_of_task, n_id, C,
                  max_iter = 200, tol = 1e-7) {
    K <- ncol(X)
    set_ll <- function(betas, pi_c) {          # eq-lc-loglik given parameters
        ll_nc <- sapply(1:C, function(c)
            persll_mnl(betas[c, ], y, X, task_id, id_of_task))
        M <- apply(ll_nc, 1, max)
        list(ll = sum(M + log(exp(ll_nc - M) %*% pi_c)), ll_nc = ll_nc)
    }
    ## start: perturbed pooled (one-class) coefficients
    pooled <- optim(rep(0, K), loglik_mnl_w, gr = grad_mnl_w, y = y, X = X,
                    task_id = task_id, w_task = rep(1, max(task_id)),
                    method = "BFGS", control = list(fnscale = -1))$par
    betas <- t(sapply(1:C, function(c) pooled + rnorm(K, 0, 0.5)))
    pi_c  <- rep(1/C, C)

    trace_ll <- numeric(0)
    for (it in 1:max_iter) {
        cur <- set_ll(betas, pi_c)
        trace_ll <- c(trace_ll, cur$ll)
        ## E-step: posterior memberships (rowwise, max-shifted)
        lw <- sweep(cur$ll_nc, 2, log(pi_c), `+`)
        lw <- lw - apply(lw, 1, max)
        w  <- exp(lw) / rowSums(exp(lw))                  # N x C
        ## M-step: weighted MNL per class; person weights mapped to tasks
        for (c in 1:C) {
            betas[c, ] <- optim(betas[c, ], loglik_mnl_w, gr = grad_mnl_w,
                                y = y, X = X, task_id = task_id,
                                w_task = w[id_of_task, c],
                                method = "BFGS",
                                control = list(fnscale = -1))$par
        }
        pi_c <- colMeans(w)
        if (it > 1 && abs(diff(tail(trace_ll, 2))) < tol) break
    }
    list(betas = betas, pi = pi_c, w = w, trace = trace_ll)
}
```

```{r}
set.seed(112)
fit_em <- em_lc(y, X, task_id, id_of_task, n_id, C = 2)
length(fit_em$trace)                        # iterations used
round(fit_em$pi, 3)
round(fit_em$betas, 2)
```

```{r}
#| label: fig-em-trace
#| fig-cap: "The EM algorithm's log-likelihood trace: monotone ascent, steep then saturating, converging to the same summit that direct BFGS maximization found."
#| fig-width: 6
#| fig-height: 3
ggplot(data.frame(iter = seq_along(fit_em$trace), ll = fit_em$trace),
       aes(iter, ll)) +
    geom_line() + geom_point(size = 0.8) +
    labs(x = "EM iteration", y = "log-likelihood")
```

And the cross-validation that replaces a package check this chapter:[^2] two *algorithmically unrelated* routes --- quasi-Newton ascent on the raw mixture likelihood versus alternating expectation and weighted fitting --- should meet at the same summit:

```{r}
c(direct_bfgs = fit_direct$value, em = max(fit_em$trace))
```

They do, to several decimals (EM's classes may arrive label-switched relative to BFGS's --- compare after ordering by price coefficient).

[^2]: The natural package validation target, `gmnl` [@sarrias2017], implements exactly this model (`model = "lc"`), but its current CRAN version is incompatible with the modern `mlogit` data layer and fails even on its own documentation example in a current R session --- a useful reminder that unmaintained packages decay. The two-independent-algorithms check used here is, if anything, the stronger discipline.

## Choosing the Number of Classes

Reality never announces $C$. The standard procedure fits a ladder of models and compares information criteria --- BIC (@sec-mnl-mle) being the conventional choice for mixtures, its stiffer penalty guarding against the mixture's appetite for spurious classes:

One convention to fix before computing: BIC's log-sample-size term uses $N = 500$ --- the number of *independent units*, people, not the $4{,}000$ tasks. The whole point of this chapter is that tasks within a person are not independent.

```{r}
#| cache: false
set.seed(113)
bic_ladder <- sapply(1:3, function(C) {
    if (C == 1) {
        f <- optim(rep(0, 6), loglik_mnl_w, gr = grad_mnl_w, y = y, X = X,
                   task_id = task_id, w_task = rep(1, max(task_id)),
                   method = "BFGS", control = list(fnscale = -1))
        ll <- f$value
    } else {
        ll <- max(replicate(3, max(em_lc(y, X, task_id, id_of_task,
                                          n_id, C)$trace)))
    }
    k <- (C - 1) + C * 6
    c(C = C, ll = ll, bic = -2*ll + k * log(n_id))
})
round(t(bic_ladder), 1)
```

(Note the `replicate(3, ...)`: each ladder rung gets multiple EM starts, per the standing local-maxima discipline.) BIC drops sharply from one class to two --- the data insist on heterogeneity --- and rises again at three: the third class buys too little fit for its six-plus parameters. The criterion recovers the truth we built in. With real data, expect less clarity, and remember what $C$ is: an instrument for approximating a heterogeneity distribution, not an ontological claim that exactly two kinds of people walk the earth. If segments are the deliverable, latent class is the model; if the heterogeneity *distribution* is the deliverable, the discrete approximation with small $C$ is crude --- the observation that opens @sec-mixed-msl.

## The Payoff: Posterior Classification

The E-step's weights are not algorithmic scaffolding; they are the model's answer to a question clients actually ask: *which segment is this respondent in?* For each person, $w_{nc}$ (@eq-estep) updates the population shares by their observed behavior --- eight consistent choices are plenty to sharpen a 60/40 prior into near-certainty:

```{r}
#| label: fig-posterior-membership
#| fig-cap: "Posterior probability of membership in class 1 for all 500 respondents. Most respondents are classified nearly decisively; the thin middle band is respondents whose eight choices genuinely fit both stories."
#| fig-width: 6
#| fig-height: 3
ggplot(data.frame(w1 = fit_em$w[, 1]), aes(w1)) +
    geom_histogram(bins = 40, fill = "grey70") +
    labs(x = "P(class 1 | choices)", y = "respondents")
```

And because this is simulated data, we can grade the classifications against the hidden truth --- the attribute smuggled out of the simulator:

```{r}
map_class <- apply(fit_em$w, 1, which.max)
truth_cls <- attr(lc, "class_assign")
## align labels (EM may have switched them), then score
acc <- max(mean(map_class == truth_cls), mean(3 - map_class == truth_cls))
round(acc, 3)
```

Roughly `r round(100*max(mean(apply(fit_em$w,1,which.max)==attr(lc,"class_assign")), mean(3-apply(fit_em$w,1,which.max)==attr(lc,"class_assign"))))`% of respondents correctly assigned from eight choices each --- with the errors concentrated, as @fig-posterior-membership predicts, among the genuinely ambiguous. This person-level inference-from-behavior is the book's first taste of its ultimate deliverable; the mixed logit's continuous version (@sec-limits) and the HB posteriors of @sec-hb-mnl are the same idea with the discrete classes melted into a continuum.

## Key Learnings

- The latent class MNL is a finite mixture: shares $\pi_c$ and class coefficients $\bfbeta_c$, with membership unobserved. Population-level IIA breaks automatically --- heterogeneity generates flexible substitution without specifying nests.
- **The mixture sits outside the product over tasks** (@eq-lc-person): people keep their class across their whole sequence, and within-person consistency is what identifies the classes. This likelihood hierarchy transfers verbatim to the mixed logit and HB models.
- The likelihood is non-concave: label switching (benign; fix by reporting convention) and genuine local maxima (not benign; fix by multiple dispersed starts --- part of the estimator, not garnish). We watched two starts reach two different summits.
- Two routes to the MLE: direct quasi-Newton on the reparameterized mixture likelihood (softmax shares), and EM --- Bayes-rule E-step (@eq-estep), weighted-MNL M-step, monotone log-likelihood ascent. Their agreement at the summit is a two-algorithm validation, stronger than any package check (and the natural package, `gmnl`, has decayed --- caveat emptor).
- The E-step weight plumbing --- person-level weights mapped to task-level likelihood terms via `w[id_of_task, c]` --- is the chapter's classic bug site: wrong-level indexing fails silently. Watch for the same hazard in every per-person model ahead.
- BIC over a ladder of $C$ values (with multiple starts per rung) recovered the true two classes; treat $C$ as an approximation device. Posterior memberships classified ~90% of respondents correctly from eight choices --- person-level inference, the book's destination, in its discrete debut.
