::: {.content-hidden}
{{< include ../latex-header.tex >}}
:::

# The Limits of Likelihood {#sec-limits}

```{r}
#| label: setup
#| message: false
#| echo: false
library(ggplot2)
theme_set(theme_minimal())
```

This short chapter is a hinge. Part II climbed as far as maximum likelihood reaches: the mixed MNL of @sec-mixed-msl estimates an entire *population* of preferences --- means, spreads, correlations --- and does it well. But look carefully at what it delivers. Population parameters. Not *people*.

For many research questions, population parameters are the answer. For the questions that dominate applied choice modeling, they are not. A firm running the laptop study doesn't stop at "price sensitivity has mean $-1.2$ and standard deviation $0.7$"; it wants to know *which respondents* are the price-insensitive premium seekers, to size that segment's demand for a new configuration, to simulate a market where each simulated buyer behaves like a particular real one (@sec-postestimation). Personalization, targeting, segment-level demand --- the deliverables of modern conjoint practice [@allenby2019] --- all need estimates of the individual $\bfbeta_n$'s. This chapter asks whether maximum likelihood can produce them. The answer --- *almost, awkwardly, without honest uncertainty* --- is the reason Part III exists.

## Attempt One: A Likelihood Per Person

The direct approach: every respondent gave us eight choices; eight choices carry information about that person's preferences; so run the @sec-mnl-mle estimator *separately on each person's data*. No population model at all --- just 500 small MNL estimations.

Load the mixed-logit study saved in @sec-mixed-msl (data, true individual coefficients, and the MSL estimates of the population parameters --- everything this chapter needs):

```{r}
mixed     <- readRDS("../data/mixed_study.rds")
d         <- mixed$data
betas_true <- mixed$betas_true          # the answer key: true beta_n, N x K
X <- model.matrix(~ brand + ram + screen + price, data = d)[, -1]
colnames(betas_true) <- colnames(X)
y <- d$choice
task_id <- cumsum(!duplicated(d[, c("id", "task")]))
id <- d$id
K <- ncol(X)
```

```{r}
loglik_mnl <- function(beta, y, X, task_id) {     # @sec-mnl-mle, unchanged
    V    <- as.vector(X %*% beta)
    Vmax <- ave(V, task_id, FUN = max)
    eV   <- exp(V - Vmax)
    sum((V - Vmax - log(rowsum(eV, task_id)[task_id]))[y == 1])
}

fit_one_person <- function(n) {
    rows <- which(id == n)
    tid  <- task_id[rows] - min(task_id[rows]) + 1
    optim(rep(0, K), loglik_mnl, y = y[rows], X = X[rows, ], task_id = tid,
          method = "BFGS", control = list(fnscale = -1, maxit = 200))$par
}

per_person <- t(sapply(1:500, fit_one_person))
colnames(per_person) <- colnames(X)
```

Five hundred estimations later, compare the per-person price coefficients to the truth we smuggled out of the simulator:

```{r}
summary_stats <- rbind(
    truth    = summary(betas_true[, "price"]),
    per_person = summary(per_person[, "price"])
)
round(summary_stats, 1)
```

The true price coefficients span a civilized range around $-1.2$. The per-person estimates span *tens or hundreds* in both directions. This is not a bug in our code; it is the estimator working as designed on a task it cannot do. Eight choices among three alternatives simply do not contain enough information to pin down six utility parameters, and the failure has two textbook faces:

- **Perfect separation.** A respondent who happened to always choose the cheapest laptop is *perfectly explained* by $\beta_{\text{price}} = -\infty$: their eight-task likelihood increases monotonically as the coefficient runs away, and `optim()` chases it until stopped by the iteration cap. With $T = 8$, such respondents are not rare accidents; they are a sizable fraction of any sample.
- **The incidental parameters problem.** Even without separation, there is a deeper pathology: when the number of parameters grows with the number of observations (six new parameters per respondent), the usual asymptotic guarantees of @sec-likelihood never engage. More respondents give you more *parameters to estimate*, not more information per parameter; $T$ is what would need to grow, and $T$ is fixed by the survey. Per-person MLE is inconsistent in the dimension that matters.[^1]

[^1]: The incidental parameters problem dates to Neyman and Scott (1948); @train2009 Section 11.4 discusses it in exactly our setting. It is why no serious practitioner estimates "fixed effects" utility coefficients from short choice panels.

```{r}
#| label: fig-per-person-fail
#| fig-cap: "Per-person MLE price coefficients (clipped to ±10 for display) against the true distribution of price coefficients. Eight choices per person cannot support six free parameters: mass piles at extreme values as separation sends estimates toward infinity."
#| fig-width: 6
#| fig-height: 3.2
df_plot <- rbind(
    data.frame(beta = pmax(pmin(per_person[, "price"], 10), -10),
               which = "per-person MLE"),
    data.frame(beta = betas_true[, "price"], which = "truth")
)
ggplot(df_plot, aes(beta, fill = which)) +
    geom_histogram(bins = 60, position = "identity", alpha = 0.6) +
    scale_fill_manual(values = c("grey30", "grey70")) +
    labs(x = expression(beta[price]), y = "count", fill = NULL)
```

Verdict: treating each person's coefficients as free parameters is a dead end --- not "needs more data" but *structurally* dead, because the survey's $T$ will never be large. Whatever produces individual estimates must let respondents share information.

## Attempt Two: Condition on the Population

Which is exactly what the mixed logit already does --- it just forgets to tell us about individuals. Recall its structure: $\bfbeta_n$ drawn from a population distribution $f(\bfbeta \mid \bftheta)$, then choices from an MNL given $\bfbeta_n$. The MSL estimate $\hat{\bftheta}$ describes the population. But the model contains a sharper question we haven't asked: *given* that person $n$ made the specific choices $\y_n$, which parts of the population distribution are they likely to sit in?

The answer is a conditional density, obtained --- as in the E-step of @sec-latentclass, and this is no coincidence --- by Bayes' rule [@train2009, Ch. 11]:

$$
h(\bfbeta \mid \y_n, \hat{\bftheta})
= \frac{ \Pr(\y_n \mid \bfbeta) \, f(\bfbeta \mid \hat{\bftheta}) }{ \int \Pr(\y_n \mid \bfbeta') \, f(\bfbeta' \mid \hat{\bftheta}) \, d\bfbeta' }
$$ {#eq-conditional-dist}

Read it with @sec-latentclass eyes: $f(\bfbeta \mid \hat{\bftheta})$ plays the role the class shares played (what people are like in general); the sequence probability $\Pr(\y_n \mid \bfbeta)$ plays the class-specific likelihood (how well each candidate taste explains *this* person); the result upgrades the population distribution into a person-specific one. The continuous analogue of posterior class membership.

Computing it is a two-line addition to machinery we already ran. The MSL estimation of @sec-mixed-msl already evaluates, for every person and every draw $\bfbeta_n^{(r)}$ from the estimated population, the sequence probability $P_{nr} = \Pr(\y_n \mid \bfbeta_n^{(r)})$ --- note the subscripts: unlike the task-level $P_{nt}$, this is the probability of person $n$'s *whole sequence* under draw $r$, i.e. $P_{nr} = e^{\ell_{nr}}$ in @sec-mixed-msl's notation, and the weights $P_{nr}/\sum_{r'} P_{nr'}$ we are about to compute are *literally* the $w_{nr}$ of @eq-msl-grad --- the Bayes-rule object that chapter promised this one would pick up. The **conditional mean** of @eq-conditional-dist follows by a move we already own. Write it out:
$$\mathrm{E}[\bfbeta \mid \y_n] = \int \bfbeta \, h(\bfbeta \mid \y_n, \hat{\bftheta}) \, d\bfbeta = \frac{\int \bfbeta \, \Pr(\y_n \mid \bfbeta) \, f(\bfbeta \mid \hat{\bftheta}) \, d\bfbeta}{\int \Pr(\y_n \mid \bfbeta') \, f(\bfbeta' \mid \hat{\bftheta}) \, d\bfbeta'}$$
Numerator and denominator are both expectations *with respect to $f$* --- exactly the kind of integral @sec-msl taught us to simulate: draw from $f$, average the integrand. Simulate both with the same draws $\bfbeta_n^{(r)} \sim f(\cdot \mid \hat{\bftheta})$, cancel the common $1/R$, and the conditional mean becomes a *probability-weighted average of the draws* --- each candidate taste weighted by how well it explains the person:

$$
\hat{\bfbeta}_n = \frac{\sum_r P_{nr} \, \bfbeta_n^{(r)}}{\sum_r P_{nr}}
$$ {#eq-conditional-mean}

```{r}
## rebuild the estimated-population draws and their weights, as in @sec-mixed-msl
theta_hat <- mixed$theta_hat            # MSL estimates: means, log-sds
rc        <- mixed$rc                    # which coefficients are random
Z         <- mixed$Z                     # the fixed draws used in estimation
R <- dim(Z)[2]; n_id <- max(id)
id_of_task <- id[!duplicated(task_id)]

bar <- theta_hat[1:K]
s   <- exp(theta_hat[(K+1):(K+length(rc))])

lseq <- matrix(0, n_id, R)
beta_draws <- array(rep(bar, each = n_id), c(n_id, R, K))
for (r in 1:R) {
    betas <- matrix(rep(bar, each = n_id), n_id, K)
    for (jj in seq_along(rc))
        betas[, rc[jj]] <- betas[, rc[jj]] + s[jj] * Z[, r, jj]
    beta_draws[, r, ] <- betas
    V    <- rowSums(X * betas[id, ])
    Vmax <- ave(V, task_id, FUN = max)
    eV   <- exp(V - Vmax)
    logP <- (V - Vmax - log(rowsum(eV, task_id)[task_id]))
    lseq[, r] <- as.vector(rowsum(logP[y == 1], id_of_task))
}
M <- apply(lseq, 1, max)
w <- exp(lseq - M); w <- w / rowSums(w)          # person-by-draw weights

beta_cond <- sapply(1:K, function(k) rowSums(w * beta_draws[, , k]))
colnames(beta_cond) <- colnames(X)
```

The moment of judgment --- conditional means against the true individual coefficients:

```{r}
#| label: fig-conditional-vs-truth
#| fig-cap: "Conditional individual-level price coefficients against each respondent's true price coefficient. Correlation is strong; the systematic flattening relative to the 45-degree line is shrinkage --- with eight tasks each, individual estimates rationally lean toward the population mean."
#| fig-width: 5.5
#| fig-height: 4
ggplot(data.frame(truth = betas_true[, "price"],
                  cond  = beta_cond[, "price"]),
       aes(truth, cond)) +
    geom_abline(linetype = "dotted") +
    geom_point(alpha = 0.35, size = 0.9) +
    coord_equal() +
    labs(x = expression(true~beta[price]),
         y = expression(conditional~mean~hat(beta)[price]))
```

```{r}
round(c(correlation = cor(betas_true[, "price"], beta_cond[, "price"]),
        sd_truth = sd(betas_true[, "price"]),
        sd_conditional = sd(beta_cond[, "price"])), 2)
```

Night and day against Attempt One. The conditional means track the truth person by person (correlation around `r round(cor(betas_true[,"price"], beta_cond[,"price"]), 2)`), with no explosions --- the population density in @eq-conditional-dist acts as a leash that separation cannot outrun: a respondent who always chose cheapest is now assigned a price coefficient in the *plausible lower tail* of the population, not at infinity.

And notice the figure's second message: the cloud is *flatter* than the 45-degree line, and the conditional means' spread is visibly smaller than the true spread. This compression is **shrinkage** --- individual estimates pulled toward the population mean, more strongly the less information the person's own choices carry. Far from a defect, shrinkage is the rational response to $T = 8$: with thin personal data, the population is a better guide than the person's noisy record, and the weighted average in @eq-conditional-mean makes precisely that trade.[^2] Shrinkage is about to become a central character: the hierarchical models of Part III produce it automatically, and @sec-hb-mnl will lay this chapter's estimates alongside theirs.

[^2]: The compression also means conditional means should *not* be treated as draws from the taste distribution --- their variance understates the population's by construction. Population statements come from $\hat{\bftheta}$; person statements from @eq-conditional-mean; mixing up the two is a standard applied error [@train2009, Section 11.5].

## So What's Missing?

If conditional means work this well, why does this book have three more parts? Because of what surrounds the point estimates. Line up the honest accounting:

**No finite-sample inference.** The conditional mean is a point. Its distribution @eq-conditional-dist would seem to offer more --- but it is computed *at the plug-in* $\hat{\bftheta}$, as if the population parameters were known rather than estimated. Person-level uncertainty statements that account for population-level uncertainty require nesting the whole @eq-conditional-mean computation inside the sampling distribution of $\hat{\bftheta}$ --- possible in principle, done by almost no one, offered by no standard software.

**A two-step architecture with error at both steps.** Step one: MSL, with its simulation bias (@sec-msl) and its own standard errors. Step two: conditional means, with fresh simulation error, treating step one as exact. The errors do not add up in any package's output; they are simply not accounted.

**And the first step itself strains.** The @sec-mixed-msl accounting: minutes per fit at three random coefficients with diagonal covariance; full covariance means Cholesky parameterizations, quadratically growing parameter counts, numerical gradients or heavy derivations, multiplied by the multiple-starts discipline. MSL *works* --- we validated it --- but every increment of model richness is paid for in optimization fragility and draws.

Now step back and look at the *shape* of what we built, because the shape is the punchline. To get individual estimates out of maximum likelihood, we bolted Bayes' rule onto the side of the estimator: @eq-conditional-dist treats the population distribution as a prior and updates it with a person's data. The natural question --- the one Part III answers --- is: *if Bayes' rule is doing the interesting work anyway, why not build the entire estimator out of it?* Treat $\bftheta$ itself as uncertain, put the hierarchy $\y_n \leftarrow \bfbeta_n \leftarrow \bftheta$ inside one coherent posterior, and sample from that posterior directly. Every deficiency in the list above dissolves: person-level and population-level uncertainty arrive together, exactly, in finite samples; there is no two-step; and --- the surprise of @sec-hb-mnl --- the computation gets *easier*, replacing a strained 9-parameter optimization with draws from simple conditional distributions. The tools are ready: the likelihood code of Part II is the engine, and the sampling skills of @sec-draws are the fuel. What remains is the algorithm --- Markov chain Monte Carlo --- and that is Part III.

## Key Learnings

- Population parameters are not the deliverable; individual-level $\bfbeta_n$'s are what targeting, segmentation, and market simulation consume. Maximum likelihood's route to them is the subject of this chapter --- and its limit.
- Per-person MLE fails structurally: perfect separation sends coefficients to infinity, and the incidental parameters problem (parameters growing with $N$, information growing only with fixed $T$) voids the asymptotics. The failure histogram is not subtle.
- Conditioning works: Bayes' rule (@eq-conditional-dist) upgrades the estimated population distribution into person-specific distributions, and the conditional mean (@eq-conditional-mean) is a probability-weighted average of draws --- computable from quantities MSL already produced, the continuous sibling of @sec-latentclass's posterior memberships.
- Conditional means track the truth strongly and exhibit **shrinkage** --- rational compression toward the population mean under thin personal data. Never confuse the spread of conditional means with the spread of tastes.
- What likelihood cannot supply: honest person-level uncertainty (the plug-in ignores $\hat{\bftheta}$'s error), a one-step architecture, and relief from MSL's growing strain. The fix is to promote Bayes' rule from bolt-on to foundation --- Part III.
