::: {.content-hidden}
{{< include ../latex-header.tex >}}
:::

# Gibbs Sampling and Hierarchical Models {#sec-gibbs}

```{r}
#| label: setup
#| message: false
#| echo: false
library(ggplot2)
theme_set(theme_minimal())
```

Metropolis--Hastings will sample anything, slowly. This chapter builds its efficient sibling and the model architecture that Part III exists for. The **Gibbs sampler** replaces M-H's propose-and-mostly-reject shuffle with draws from *exact conditional distributions* --- proposals so good they are never rejected --- whenever the model's structure supplies such conditionals. And the model structure that supplies them most beautifully is the **hierarchy**: person-level parameters drawn from a population, the formal home of everything @sec-limits found missing. We develop both on models simple enough to see through --- a hierarchical normal model, then a binary probit --- because @sec-hb-mnl assembles the HB-MNL *entirely* from blocks built here, and @sec-mnp-est's probit samplers are these ideas at scale.

## Sampling by Blocks

Suppose the parameter vector splits into blocks $\bftheta = (\bftheta_1, \bftheta_2, \ldots)$ and, while the *joint* posterior is intractable, each **full conditional** --- the distribution of one block given the data and all other blocks --- is a known, sampleable distribution. The Gibbs sampler simply cycles:

$$
\bftheta_1^{(s+1)} \sim p(\bftheta_1 \mid \bftheta_2^{(s)}, \y), \qquad
\bftheta_2^{(s+1)} \sim p(\bftheta_2 \mid \bftheta_1^{(s+1)}, \y), \qquad \ldots
$$ {#eq-gibbs-cycle}

--- always conditioning on the most recent values of the other blocks. The resulting chain has the joint posterior as its stationary distribution, exactly like M-H;[^1] the practical differences are that every draw is accepted, there is no step size to tune, and the moves are calibrated to the local shape of the posterior *by construction*. The cost of admission: you must be able to derive and sample the conditionals --- real analytic work, purchased model by model, mostly through a phenomenon called conjugacy.

[^1]: The connection is exact: a Gibbs block-update is a Metropolis--Hastings step whose proposal is the full conditional itself, for which the acceptance ratio works out to precisely 1. One algorithm, two ends of a proposal-quality spectrum --- a fact we exploit in @sec-hb-mnl by mixing the two within one cycle.

**A toy first**, per the @sec-bayes-mnl discipline of testing samplers on known targets. The bivariate normal with correlation $\rho$ has textbook conditionals --- each coordinate given the other is univariate normal, $x_1 | x_2 \sim N(\rho x_2, 1 - \rho^2)$ and symmetrically --- so Gibbs needs six lines, and @sec-mvnorm's Cholesky method provides an exact benchmark:

```{r}
set.seed(160)
rho <- 0.8; S <- 10000
gx <- matrix(NA_real_, S, 2); xy <- c(-3, 3)         # deliberately absurd start
for (s in 1:S) {
    xy[1] <- rnorm(1, rho * xy[2], sqrt(1 - rho^2))
    xy[2] <- rnorm(1, rho * xy[1], sqrt(1 - rho^2))
    gx[s, ] <- xy
}
rbind(gibbs   = c(mean = colMeans(gx[-(1:100), ]),
                  cor = cor(gx[-(1:100), 1], gx[-(1:100), 2])),
      target  = c(0, 0, rho))
```

Means and correlation on target. The chain's *path* is worth one look --- Gibbs moves in axis-parallel steps (one coordinate at a time), staircasing across the correlated ellipse:

```{r}
#| label: fig-gibbs-staircase
#| fig-cap: "The first 30 moves of the Gibbs sampler on a correlated bivariate normal, from a far-off start: axis-parallel staircase steps, because each update changes one coordinate conditional on the other. High correlation makes the staircase treads small -- a preview of why strongly dependent parameters mix slowly under coordinate-wise sampling."
#| fig-width: 5
#| fig-height: 4
path <- data.frame(x1 = rep(gx[1:30, 1], each = 2)[-1],
                   x2 = c(3, rep(gx[1:30, 2], each = 2))[1:59])
ggplot(path, aes(x1, x2)) +
    geom_path(linewidth = 0.3) +
    geom_point(data = data.frame(x1 = gx[1:30, 1], x2 = gx[1:30, 2]),
               size = 0.8) +
    coord_equal() +
    labs(x = expression(x[1]), y = expression(x[2]))
```

Note the figure's caption warning: correlation between blocks shrinks the staircase treads. It will return as practical advice --- block correlated parameters *together* when you can.

## Conjugacy: The Conditional-Distribution Factory

Where do sampleable conditionals come from? From prior-likelihood pairs whose product lands back in the prior's family --- **conjugate** pairs. One derivation, done honestly, powers everything below; the rest we state and use.

**The normal--normal update.** Data $y_1, \ldots, y_n \iid N(\theta, \sigma^2)$ with $\sigma^2$ known; prior $\theta \sim N(\mu_0, \tau_0^2)$. The posterior is proportional to prior times likelihood --- two exponentials of quadratics in $\theta$:
$$p(\theta \mid \y) \;\propto\; \exp\!\left( -\frac{(\theta - \mu_0)^2}{2\tau_0^2} \right) \prod_{i=1}^{n} \exp\!\left( -\frac{(y_i - \theta)^2}{2\sigma^2} \right) \;\propto\; \exp\!\left( -\tfrac{1}{2}\!\left[ \theta^2 \left( \tau_0^{-2} + n\sigma^{-2} \right) - 2\theta \left( \tau_0^{-2}\mu_0 + n\sigma^{-2}\bar{y} \right) \right] \right)$$
where every term free of $\theta$ has been dropped into the proportionality sign. Now a reading-off rule worth memorizing on its own: anything of the form $\exp(-\tfrac{1}{2}[a\theta^2 - 2b\theta])$ is an unnormalized $N(b/a, \; 1/a)$ density --- complete the square to see it. With $a = \tau_0^{-2} + n\sigma^{-2}$ and $b = \tau_0^{-2}\mu_0 + n\sigma^{-2}\bar{y}$, we read off

$$
\theta \mid \y \;\sim\; N\!\left( \frac{ \tau_0^{-2} \, \mu_0 + n\sigma^{-2} \, \bar{y} }{ \tau_0^{-2} + n\sigma^{-2} }, \;\; \left( \tau_0^{-2} + n\sigma^{-2} \right)^{-1} \right)
$$ {#eq-normal-normal}

Memorize the *shape*, not the symbols: **precisions add; the posterior mean is the precision-weighted average of prior mean and data mean.** Rich data (large $n\sigma^{-2}$) pull the posterior onto $\bar{y}$; thin data leave it near $\mu_0$. That precision-weighted compromise *is* shrinkage --- @sec-limits's phenomenon, now with an exact formula --- and you will see this update, in scalar and matrix dress, inside every sampler for the rest of the book.

**Variances: the inverse-gamma update.** For the variance of a normal with known mean, the conjugate prior is inverse-gamma: $\sigma^2 \sim IG(a_0, b_0)$ yields $\sigma^2 \mid \y \sim IG(a_0 + n/2, \; b_0 + \tfrac{1}{2}\sum (y_i - \theta)^2)$ --- prior pseudo-observations plus data observations, prior sum-of-squares plus data sum-of-squares. (Its matrix generalization, the inverse-Wishart for covariance matrices, arrives in @sec-hb-mnl; same grammar.) Derivations: @bda3, or @rossi2005 in exactly our setting.

## The Hierarchical Normal Model

Now the architecture. Strip the choice modeling away from @sec-limits's problem and what remains is this: repeated measurements on many units, each unit with its own parameter, the parameters themselves drawn from a population we also want to learn. The minimal version --- our laptop shoppers, with each person's *willingness-to-spend on upgrades* $\beta_n$ measured noisily $T$ times:[^2]

$$
y_{nt} \sim N(\beta_n, \sigma^2), \qquad
\beta_n \sim N(\mu, \tau^2), \qquad
\mu \sim N(0, 10^2), \;\; \tau^2 \sim IG(2, 2), \;\; \sigma^2 \sim IG(2, 2)
$$ {#eq-hier-normal}

(A notation caution: this $\tau^2$ is the population *variance* --- how much true willingness-to-spend varies across people --- and has nothing to do with @sec-bayes-mnl's Metropolis step size $\tau$. Gibbs has no step size, so the letter is free again, and $\tau$ for a hierarchy's spread is the standard choice.)

[^2]: The stripped-down story: imagine observing, for each of $N$ shoppers, $T$ noisy dollar amounts spent on laptop upgrades. Nothing here is choice modeling --- deliberately. The hierarchy is the lesson; the MNL kernel plugs back in next chapter.

Three layers: data given person-parameters, person-parameters given population, priors on everything unknown. This *is* the mixed model of @sec-mixed-msl in normal-linear miniature --- and where MSL had to integrate the $\beta_n$'s out (they were nuisance obstacles to the likelihood), the Bayesian treatment keeps every $\beta_n$ *in* the posterior as a first-class unknown: $p(\bfbeta_{1:N}, \mu, \tau^2, \sigma^2 \mid \y)$. What looked like the problem (hundreds of parameters!) becomes the solution, because every full conditional is conjugate:

- $\beta_n \mid \cdots$: normal--normal update (@eq-normal-normal) with "prior" $N(\mu, \tau^2)$ and "data" person $n$'s $T$ observations. *One precision-weighted shrinkage per person.*
- $\mu \mid \cdots$: normal--normal update treating the current $\beta_n$'s as data.
- $\tau^2 \mid \cdots$ and $\sigma^2 \mid \cdots$ (where "$\cdots$" always means *the data and every other current block*): inverse-gamma updates, each applied to the residuals its block owns. The $\beta_n$'s are $\tau^2$'s "data": $\tau^2 \mid \cdots \sim IG\big(a_0 + N/2, \; b_0 + \tfrac{1}{2}\sum_n (\beta_n - \mu)^2\big)$. The observations are $\sigma^2$'s: $\sigma^2 \mid \cdots \sim IG\big(a_0 + NT/2, \; b_0 + \tfrac{1}{2}\sum_{n,t} (y_{nt} - \beta_n)^2\big)$.

```{r}
gibbs_hier <- function(y, id, S, mu0 = 0, tau0sq = 100, a0 = 2, b0 = 2) {
    N  <- max(id)
    Tn <- tabulate(id)                    # obs per person
    ybar_n <- as.vector(rowsum(y, id)) / Tn

    beta <- ybar_n; mu <- mean(y); tau2 <- 1; sig2 <- 1
    out <- list(beta = matrix(NA_real_, S, N), hyper = matrix(NA_real_, S, 3))
    for (s in 1:S) {
        ## person-level betas: precision-weighted normal draws, all N at once
        prec <- 1/tau2 + Tn/sig2
        m    <- (mu/tau2 + Tn*ybar_n/sig2) / prec
        beta <- rnorm(N, m, sqrt(1/prec))
        ## population mean
        precm <- 1/tau0sq + N/tau2
        mu    <- rnorm(1, (mu0/tau0sq + sum(beta)/tau2) / precm, sqrt(1/precm))
        ## population variance and noise variance
        tau2 <- 1 / rgamma(1, a0 + N/2, b0 + sum((beta - mu)^2)/2)
        sig2 <- 1 / rgamma(1, a0 + length(y)/2, b0 + sum((y - beta[id])^2)/2)
        out$beta[s, ]  <- beta
        out$hyper[s, ] <- c(mu, sqrt(tau2), sqrt(sig2))
    }
    colnames(out$hyper) <- c("mu", "tau", "sigma")
    out
}
```

Every line is one of the two conjugate updates. Note the vectorization: all $N$ person-draws happen in a single `rnorm(N, ...)` call, because given the hyperparameters the $\beta_n$'s are conditionally *independent* --- the hierarchy's structure is the code's structure (and `matrix(NA_real_, S, N)` is @sec-performance's preallocation rule earning its keep: this sampler stores half a million numbers). Simulate and recover:

```{r}
set.seed(161)
N <- 200; T <- 5
mu_true <- 2; tau_true <- 1.5; sig_true <- 2
beta_n <- rnorm(N, mu_true, tau_true)
id <- rep(1:N, each = T)
y  <- rnorm(N*T, beta_n[id], sig_true)

fit <- gibbs_hier(y, id, S = 5000)
post <- fit$hyper[-(1:1000), ]
round(rbind(truth = c(mu_true, tau_true, sig_true),
            post_mean = colMeans(post),
            post_sd   = apply(post, 2, sd)), 2)
```

All three population parameters recovered, uncertainty attached, in seconds --- no optimizer, no draws-inside-a-likelihood, no tuning. And the person level came along for free. Here is this book's first *fully honest* individual-level inference, and the shrinkage geometry it produces:

```{r}
#| label: fig-shrinkage
#| fig-cap: "Shrinkage, exactly: each person's posterior-mean beta against their raw data mean (T = 5 observations each). The cloud tilts flatter than the 45-degree line, pulled toward the population mean (dashed) -- extreme raw means are partly noise, and the precision-weighted update (@eq-normal-normal) discounts them accordingly."
#| fig-width: 5.5
#| fig-height: 4
ybar_n <- as.vector(rowsum(y, id)) / T
df <- data.frame(raw = ybar_n, post = colMeans(fit$beta[-(1:1000), ]))
ggplot(df, aes(raw, post)) +
    geom_abline(linetype = "dotted") +
    geom_hline(yintercept = mean(post[, "mu"]), linetype = "dashed") +
    geom_point(alpha = 0.4, size = 0.9) +
    coord_equal() +
    labs(x = "person's raw mean", y = "person's posterior mean")
```

The flattened cloud is @eq-normal-normal acting on two hundred people at once: each posterior mean is a precision-weighted compromise between that person's own average and the population's. @sec-limits met this as an ad-hoc extraction; here it is a *theorem-grade consequence of the model*, with the amount of shrinkage --- the balance of $\tau^{-2}$ against $T\sigma^{-2}$ --- itself estimated from the data. More data per person, less shrinkage; more real population spread, less shrinkage; the model negotiates it, not the analyst. This figure is the entire promise of Part III in one frame; @sec-hb-mnl reproduces it with choice data.

## Data Augmentation: The Albert--Chib Sampler

The hierarchy needed no new trick --- conjugacy carried it. Choice data will, because a *logit or probit likelihood is not conjugate to anything*: the discreteness of $y$ breaks the quadratic-in-parameters structure that @eq-normal-normal feeds on. The escape, due to @albertchib1993, is one of the loveliest moves in computational statistics: *put the latent utilities back.*

Recall the binary probit of @sec-simdata: $y_n = I[z_n > 0]$ where $z_n = \x_n'\bfbeta + \varepsilon_n$, $\varepsilon_n \sim N(0,1)$. The observed $y_n$ is a censored glimpse of a well-behaved linear-normal model in $z_n$. So **augment**: treat $\mathbf{z}$ as unknowns alongside $\bfbeta$, and Gibbs the pair --- each conditional is suddenly easy:

- $\bfbeta \mid \mathbf{z}$: with $\mathbf{z}$ "observed," this is *Bayesian linear regression* --- normal prior, normal likelihood, so a multivariate @eq-normal-normal: $\bfbeta \sim N\big( (X'X + B_0^{-1})^{-1}(X'\mathbf{z} + B_0^{-1}\mathbf{b}_0), \; (X'X + B_0^{-1})^{-1} \big)$.
- $z_n \mid \bfbeta, y_n$: normal with mean $\x_n'\bfbeta$, *truncated* to the side of zero that agrees with the observed choice --- positive if they bought, negative if not. Which is exactly what @sec-draws's `rtnorm()` was built for, as promised there.

```{r}
rtnorm <- function(n, mean = 0, sd = 1, a = -Inf, b = Inf) {   # @sec-draws
    Fa <- pnorm(a, mean, sd); Fb <- pnorm(b, mean, sd)
    qnorm(Fa + runif(n) * (Fb - Fa), mean, sd)
}

gibbs_probit <- function(y, X, S, b0 = 0, B0 = 25) {
    K  <- ncol(X)
    B0inv <- diag(1/B0, K)
    V  <- solve(crossprod(X) + B0inv)              # fixed: X never changes
    L  <- t(chol(V))                                # @sec-mvnorm, once
    beta  <- rep(0, K)
    lo <- ifelse(y == 1, 0, -Inf); hi <- ifelse(y == 1, Inf, 0)
    draws <- matrix(NA_real_, S, K)
    for (s in 1:S) {
        z    <- rtnorm(length(y), as.vector(X %*% beta), 1, lo, hi)
        m    <- V %*% (crossprod(X, z) + B0inv %*% rep(b0, K))
        beta <- as.vector(m + L %*% rnorm(K))
        draws[s, ] <- beta
    }
    draws
}
```

Count what is absent: no Metropolis step, no acceptance rate, no tuning parameter, no likelihood evaluation --- the probit *likelihood never appears in the code*. Its integrals (the $\Phi$'s that a probit MLE would evaluate) have been traded for truncated-normal draws, and every iteration draws from exact distributions. Note also the @sec-performance economies: $V$ and its Cholesky factor are data-only quantities, computed once outside the loop.

Test on binary probit laptop data --- @sec-simdata's simulator, `dist = "normal"`:

```{r}
set.seed(162)
sim_binary_data <- function(N, beta, dist = "normal") {
    price <- runif(N, 0.8, 2.4); ram16 <- rbinom(N, 1, 0.5)
    X <- cbind(1, price, ram16)
    eps <- if (dist == "normal") rnorm(N) else rlogis(N)
    data.frame(y = as.integer(as.vector(X %*% beta) + eps > 0), price, ram16)
}
beta_true <- c(1.0, -1.2, 0.6)
d <- sim_binary_data(4000, beta_true)
X <- cbind(1, d$price, d$ram16)

ac <- gibbs_probit(d$y, X, S = 5000)
round(rbind(truth = beta_true,
            post_mean = colMeans(ac[-(1:500), ]),
            post_sd   = apply(ac[-(1:500), ], 2, sd)), 3)
```

Recovery, tuning-free. Data augmentation is this book's quiet superweapon: in @sec-mnp-est it makes the *multinomial* probit --- whose likelihood is a $(J-1)$-dimensional integral that Part II could only simulate --- estimable by nothing more than truncated normals and regressions, and in @sec-hb-mnp it does so with a full hierarchy on top. The general principle deserves its box: **when a likelihood is intractable because something latent was integrated out, put the latent thing back and sample it.** Latent utilities, latent classes (@sec-latentclass's memberships can be Gibbs-sampled the same way), latent budgets (@sec-extensions) --- all one move.

## Metropolis-within-Gibbs: The Missing Piece

Survey what Part III now has. M-H samples anything evaluable (@sec-bayes-mnl) but wanders. Gibbs is exact and tuning-free but demands conjugate conditionals. The hierarchy gives person-level inference with automatic shrinkage --- when the data layer is normal. And augmentation converts probit data layers into normal ones.

One target resists: the **MNL data layer**. The Gumbel-error structure has no Albert--Chib-style augmentation into conjugacy;[^3] the logit likelihood must be *evaluated*, not conjugated away. But nothing requires a Gibbs cycle to be conjugate in *every* block. Replace the unavailable exact draw, in just that block, with a Metropolis step targeting the block's conditional --- accept/reject by @eq-mh-ratio, conditioning on all other blocks --- and the chain remains valid: **Metropolis-within-Gibbs**. The blueprint for the next chapter writes itself: population mean and covariance, conjugate Gibbs draws (@eq-normal-normal and inverse-Wishart, treating the current $\bfbeta_n$'s as data --- literally `gibbs_hier`'s middle blocks); each person's $\bfbeta_n$, a small random-walk Metropolis step whose target is (population prior) × (that person's eight-task MNL likelihood) --- literally @sec-bayes-mnl's sampler, shrunk to one person's data. Every component exists. Assembly is @sec-hb-mnl.

[^3]: A Pólya-Gamma augmentation *does* exist for logit models and is a lovely modern development --- but the Metropolis-within-Gibbs route is standard in choice modeling, more general, and builds directly on our @sec-bayes-mnl machinery, so we take it.

## Key Learnings

- Gibbs sampling cycles through blocks, drawing each from its full conditional (@eq-gibbs-cycle): every draw accepted, nothing tuned, at the price of deriving conditionals. It is M-H with perfect proposals --- and correlated blocks make its staircase treads small, so block wisely.
- Conjugacy is the conditional factory. The normal--normal update (@eq-normal-normal) --- *precisions add, means precision-weight* --- is the single most reused formula in the rest of this book; inverse-gamma (and later inverse-Wishart) handles the variances.
- The hierarchical normal model holds every person's parameter in the posterior jointly with the population's: person-level inference with *estimated, automatic shrinkage* (@fig-shrinkage) --- @sec-limits's ad-hoc extraction, rebuilt as a consequence of the model.
- Data augmentation (Albert--Chib): reinstate latent utilities and the probit's intractable likelihood dissolves into truncated-normal draws (`rtnorm()`, as promised in @sec-draws) plus Bayesian regression. No tuning, no likelihood evaluations. The same move powers the MNP samplers of Part IV.
- The MNL layer alone resists conjugation; Metropolis-within-Gibbs --- exact draws where available, M-H steps where not --- closes the gap. All HB-MNL components are now on the bench.
