::: {.content-hidden}
{{< include ../latex-header.tex >}}
:::

# The Hierarchical Bayesian MNL {#sec-hb-mnl}

```{r}
#| label: setup
#| message: false
#| echo: false
library(ggplot2)
theme_set(theme_minimal())
```

This is the destination chapter. The hierarchical Bayesian multinomial logit --- **HB-MNL** to the industry that runs it thousands of times a day --- is the model this book has been assembling since page one: MNL choice behavior (@sec-mnl), person-specific coefficients (@sec-latentclass, @sec-mixed-msl), a population distribution binding them (@sec-gibbs), and a posterior explored by MCMC (@sec-bayes-mnl). Every block of its sampler is code you have already written and tested; this chapter's work is assembly, and then the payoff that @sec-limits promised: exact person-level inference, with the shrinkage negotiated by the model, at computational cost that *falls* where MSL's rose.

We estimate it on the same simulated dataset as @sec-mixed-msl and @sec-limits --- one dataset, three estimators, honest comparison.

## The Model

Three layers, written once in full:

$$
\begin{aligned}
\text{choices:} \quad & y_{nt} \sim \text{MNL}\left( \x_{nt1}'\bfbeta_n, \ldots, \x_{ntJ}'\bfbeta_n \right) \\
\text{people:} \quad & \bfbeta_n \sim N\left( \bar{\bfbeta}, \Sigma \right) \\
\text{priors:} \quad & \bar{\bfbeta} \sim N\left( \mathbf{0}, 100 \, I \right), \qquad \Sigma \sim IW\left( \nu_0, V_0 \right)
\end{aligned}
$$ {#eq-hb-mnl}

The first two layers *are* the mixed MNL of @eq-mixed-model --- identical model of behavior and heterogeneity. The third layer is what makes it Bayesian: the population parameters themselves get (weakly informative) priors, standard choices being $\nu_0 = K + 3$ degrees of freedom and scale $V_0 = \nu_0 I$ for the inverse-Wishart [@rossi2005 --- and mark that $V_0$; it returns under scrutiny in the practical notes]. Two specification notes. We let **all six** coefficients be random with a **full covariance** $\Sigma$ --- the specification that cost MSL its 27-parameter nightmare is the default here, because (as the sampler will show) the hierarchy prices covariance matrices in draws, not in optimization dimensions; it is also deliberately *more general than the truth* (which has three random coefficients, uncorrelated), letting us watch how an honest model handles generality it doesn't need. And unlike @sec-mixed-msl there is no draw count $R$, no Halton bases, no simulated integral anywhere: the $\bfbeta_n$ are not integrated out but *sampled*, which is the whole trick.

## The Sampler

The joint posterior over everything unknown --- $p(\bfbeta_{1:N}, \bar{\bfbeta}, \Sigma \mid \text{data})$, a $(500 \times 6 + 6 + 21)$-dimensional object --- is intractable whole but factors beautifully into @sec-gibbs blocks:

**Block 1: each $\bfbeta_n$.** Given $\bar{\bfbeta}$ and $\Sigma$, person $n$'s conditional posterior is (their MNL likelihood over $T=8$ tasks) $\times$ (their $N(\bar{\bfbeta}, \Sigma)$ population prior) --- exactly @eq-conditional-dist, now sampled rather than summarized. The MNL kernel blocks conjugacy (@sec-gibbs's closing diagnosis), so this block takes a **random-walk Metropolis step**: propose $\bfbeta_n' = \bfbeta_n + \tau L\mathbf{u}$ --- with $\mathbf{u} \sim N(\mathbf{0}, I_K)$ a fresh standard-normal vector --- and accept by the log ratio. Two refinements from the trade [@train2001; @rossi2005]: the proposal is shaped by the *current population covariance* ($L$ the lower-triangular Cholesky factor, $LL' = \Sigma$; in R, `t(chol(Sig))`, since `chol()` returns the upper factor --- steps stretched along directions the population actually varies), and the step size $\tau$ --- @sec-bayes-mnl's one knob, named `step` in the code --- **adapts during burn-in** toward a ~30% acceptance rate, then freezes.[^1]

Explicitly, person $n$'s move is accepted with probability
$$\alpha_n = \min\left(1, \; \frac{ \left[\prod_{t} P(y_{nt} \mid \bfbeta_n')\right] \phi(\bfbeta_n' \mid \bar{\bfbeta}, \Sigma) }{ \left[\prod_{t} P(y_{nt} \mid \bfbeta_n)\right] \phi(\bfbeta_n \mid \bar{\bfbeta}, \Sigma) } \right)$$
--- @sec-bayes-mnl's @eq-mh-ratio with the population density $\phi(\cdot \mid \bar{\bfbeta}, \Sigma)$ playing the prior's role.

[^1]: Freezing matters: a chain whose proposal keeps adapting forever is not quite a Markov chain, and its stationarity guarantees decay. Adapt during burn-in (discarded anyway), then run fixed. Also note the happy structure: given $(\bar{\bfbeta}, \Sigma)$, the $N$ person-blocks are conditionally *independent* --- so all 500 Metropolis steps can be executed simultaneously with matrix operations: propose an $N \times K$ matrix, evaluate all persons' likelihoods with one stacked @sec-mnl-mle kernel call, accept element-wise. Two kernel calls per iteration, total, for all people. That is why this sampler runs in minutes.

**Block 2: $\bar{\bfbeta}$.** Given the current $\bfbeta_n$'s and $\Sigma$, this is the @eq-normal-normal update with the $\bfbeta_n$'s as "data" --- a conjugate multivariate normal draw. Literally `gibbs_hier`'s middle block from @sec-gibbs. With the prior parameterized by its *precision* $A_0$ --- so $N(\mathbf{0}, A_0^{-1}I)$ with $A_0 = 0.01$ is exactly @eq-hb-mnl's $N(\mathbf{0}, 100I)$; precision is how the prior enters the update, and the hierarchical literature's convention (and `bayesm`'s) is to pass it directly --- precisions add exactly as in @eq-normal-normal:
$$\bar{\bfbeta} \mid \bfbeta_{1:N}, \Sigma \;\sim\; N\!\Big( \big(N\Sigma^{-1} + A_0 I\big)^{-1} \Sigma^{-1} \textstyle\sum_n \bfbeta_n, \;\; \big(N\Sigma^{-1} + A_0 I\big)^{-1} \Big)$$

**Block 3: $\Sigma$.** Given $\bfbeta_n$'s and $\bar{\bfbeta}$: inverse-Wishart draw from the accumulated deviations --- the matrix version of @sec-gibbs's variance update. Sampling an IW is one small function via the Bartlett decomposition (build a Wishart from chi-square and normal draws --- a construction in the spirit of @sec-draws --- and invert).
$$\Sigma \mid \bfbeta_{1:N}, \bar{\bfbeta} \;\sim\; IW\!\Big( \nu_0 + N, \;\; V_0 + \textstyle\sum_n (\bfbeta_n - \bar{\bfbeta})(\bfbeta_n - \bar{\bfbeta})' \Big)$$
--- prior degrees of freedom plus data observations, prior scale plus data scatter: @sec-gibbs's inverse-gamma grammar, in matrix dress.

Metropolis-within-Gibbs, exactly as @sec-gibbs forecast: two exact draws and one tuned step per cycle.

## Coding It

```{r}
study <- readRDS("../data/mixed_study.rds")
d <- study$data
X <- model.matrix(~ brand + ram + screen + price, data = d)[, -1]
y <- d$choice
task_id <- cumsum(!duplicated(d[, c("id", "task")]))
id <- d$id
n_id <- max(id); K <- ncol(X)

## per-person MNL log-likelihood, given an N x K matrix of betas
## (the @sec-mnl-mle kernel + one rowsum by person; cf. @sec-mixed-msl)
persll <- function(B, y, X, task_id, id, id_of_task) {
    V    <- rowSums(X * B[id, ])
    Vmax <- ave(V, task_id, FUN = max)
    eV   <- exp(V - Vmax)
    logP <- V - Vmax - log(rowsum(eV, task_id)[task_id])
    as.vector(rowsum(logP[y == 1], id_of_task))
}

## inverse-Wishart via the Bartlett decomposition
rinvwishart <- function(nu, S) {
    K <- nrow(S); L <- t(chol(solve(S)))
    A <- matrix(0, K, K)
    diag(A) <- sqrt(rchisq(K, nu - (1:K) + 1))
    A[lower.tri(A)] <- rnorm(K*(K-1)/2)
    solve(L %*% A %*% t(A) %*% t(L))
}
```

The only new machinery relative to @sec-mnl-mle's kernel is `persll()`'s last line: `logP[y == 1]` has one entry per task (the chosen alternative's log probability), and `id_of_task` labels each task with its owner, so `rowsum()` adds each person's eight task log-probabilities into the length-$N$ vector the Metropolis step consumes.

```{r}
hb_mnl <- function(y, X, task_id, id, S_iter, burn,
                   nu0 = ncol(X) + 3, V0 = diag(ncol(X)) * (ncol(X) + 3),
                   A0 = 0.01, step0 = 0.1, thin_ind = 10) {
    K <- ncol(X); n_id <- max(id)
    id_of_task <- id[!duplicated(task_id)]

    B    <- matrix(0, n_id, K)                 # person betas
    bar  <- rep(0, K); Sig <- diag(K)
    step <- step0
    ll_cur <- persll(B, y, X, task_id, id, id_of_task)

    out <- list(bar = matrix(NA_real_, S_iter, K),
                sd  = matrix(NA_real_, S_iter, K),
                acc = numeric(S_iter),
                B_draws = array(NA_real_, c((S_iter - burn) %/% thin_ind, n_id, K)))
    kk <- 0
    for (s in 1:S_iter) {
        ## -- block 1: all beta_n at once, RW Metropolis --
        L <- t(chol(Sig))
        prop <- B + step * (matrix(rnorm(n_id*K), n_id, K) %*% t(L))
        ll_prop <- persll(prop, y, X, task_id, id, id_of_task)
        Q_cur  <- forwardsolve(L, t(B    - rep(bar, each = n_id)))
        Q_prop <- forwardsolve(L, t(prop - rep(bar, each = n_id)))
        lpri_cur  <- -0.5 * colSums(Q_cur^2)     # log N(bar, Sig), up to const
        lpri_prop <- -0.5 * colSums(Q_prop^2)
        accept <- log(runif(n_id)) < (ll_prop + lpri_prop - ll_cur - lpri_cur)
        B[accept, ] <- prop[accept, ]
        ll_cur[accept] <- ll_prop[accept]
        if (s <= burn) step <- step * ifelse(mean(accept) > 0.30, 1.05, 0.95)

        ## -- block 2: bar | B, Sig (conjugate normal; prior N(0, (1/A0) I)) --
        Sinv <- solve(Sig)
        Vb   <- solve(n_id * Sinv + diag(A0, K))
        bar  <- as.vector(Vb %*% (Sinv %*% colSums(B)) +
                          t(chol(Vb)) %*% rnorm(K))

        ## -- block 3: Sigma | B, bar (inverse-Wishart) --
        Dev <- B - rep(bar, each = n_id)
        Sig <- rinvwishart(nu0 + n_id, V0 + crossprod(Dev))

        out$bar[s, ] <- bar
        out$sd[s, ]  <- sqrt(diag(Sig))
        out$acc[s]   <- mean(accept)
        if (s > burn && (s - burn) %% thin_ind == 0) {
            kk <- kk + 1; out$B_draws[kk, , ] <- B
        }
    }
    out
}
```

Sixty lines, and read against the blueprint every one is accounted for: the vectorized all-persons Metropolis step (two `persll()` calls per iteration --- footnote 1's promise, kept), the population prior evaluated via `forwardsolve()` against $\Sigma$'s Cholesky factor (the @sec-mvnorm quadratic form, no densities packages), the two conjugate draws, the burn-in-only step adaptation, and @sec-performance's preallocation (with the person-draws thinned --- storing every 10th of $500 \times 6$ numbers keeps memory civil at no inferential cost). One subtlety earns its own sentence: `lpri_cur` and `lpri_prop` omit the normal density's constant $-\tfrac{1}{2}\log|\Sigma| - \tfrac{K}{2}\log 2\pi$, which is legitimate only because both are evaluated under the *same* $\Sigma$ within an iteration, so the constant cancels in the acceptance ratio. Mechanically, `forwardsolve(L, ...)` computes $L^{-1}(\bfbeta_n - \bar{\bfbeta})$, whose squared column sums are the quadratic forms $(\bfbeta_n - \bar{\bfbeta})'\Sigma^{-1}(\bfbeta_n - \bar{\bfbeta})$ for all 500 people in one triangular solve. And one bookkeeping asymmetry to hold onto: `bar` and `sd` store every iteration, burn-in included --- we drop the first `burn` rows at analysis time --- while the person-level draws are kept only after burn-in and only every `thin_ind`-th cycle, so each chain's `B_draws` is a $400 \times 500 \times 6$ array: 400 thinned draws, 500 people, 6 coefficients.

## Running It

```{r}
#| cache: false
set.seed(170)
S_iter <- 5000; burn <- 1000
fit  <- hb_mnl(y, X, task_id, id, S_iter = S_iter, burn = burn)
fit2 <- hb_mnl(y, X, task_id, id, S_iter = S_iter, burn = burn)   # second chain
round(c(acc_rate = mean(fit$acc[-(1:burn)])), 2)
```

Two independent chains (different starting randomness), a minute or so each --- pause on that: the *full-covariance* hierarchical model, on data where MSL with a diagonal 3-coefficient specification took comparable time per fit, and where the full-covariance MSL was an extrapolated nightmare. Diagnostics, per @sec-bayes-mnl's non-negotiable workflow:

```{r}
#| label: fig-hb-trace
#| fig-cap: "Traces of the population mean and standard deviation of the price coefficient, two chains overlaid (post-burn-in, thinned for display). Fuzzy, overlapping, stationary: the caterpillars of @sec-bayes-mnl, now on a 3,027-parameter posterior."
#| fig-width: 6.5
#| fig-height: 3.6
post <- (burn + 1):S_iter
tr <- rbind(
    data.frame(iter = post, chain = "1", what = "bar_price",  v = fit$bar[post, 6]),
    data.frame(iter = post, chain = "2", what = "bar_price",  v = fit2$bar[post, 6]),
    data.frame(iter = post, chain = "1", what = "sd_price",   v = fit$sd[post, 6]),
    data.frame(iter = post, chain = "2", what = "sd_price",   v = fit2$sd[post, 6])
)
ggplot(tr[tr$iter %% 4 == 0, ], aes(iter, v, color = chain)) +
    geom_line(linewidth = 0.2) +
    facet_wrap(~ what, scales = "free_y", ncol = 1) +
    scale_color_grey(end = 0.6) +
    labs(x = "iteration", y = NULL)
```

```{r}
rhat <- function(mat) {                       # from @sec-bayes-mnl
    m <- ncol(mat); n <- nrow(mat)
    B <- n * var(colMeans(mat)); W <- mean(apply(mat, 2, var))
    sqrt(((n - 1)/n * W + B/n) / W)
}
round(range(sapply(1:K, function(k)
    rhat(cbind(fit$bar[post, k], fit2$bar[post, k])))), 3)
```

Healthy. Note in the traces that `sd_price` mixes more slowly than `bar_price` --- variance components always do (they are informed only by dispersion *between* noisy person-estimates), and their effective sample sizes are correspondingly lower. Population-level results, pooling the chains:

```{r}
bar_draws <- rbind(fit$bar[post, ], fit2$bar[post, ])
sd_draws  <- rbind(fit$sd[post, ],  fit2$sd[post, ])
tab <- data.frame(
    truth_bar = study$bar_true,   post_bar = colMeans(bar_draws),
    bar_sd    = apply(bar_draws, 2, sd),
    truth_sd  = study$sd_true,    post_sd  = colMeans(sd_draws)
)
rownames(tab) <- colnames(X)
round(tab, 2)
```

Read the two halves separately. The population *means* recover well, if a touch generously scaled. The population *standard deviations* recover the pattern --- the three truly-random coefficients (`apple`, `ram32`, `price`) estimated largest, in the right order --- but the three truly-*fixed* coefficients show standard deviations near 0.6 rather than 0. That is not a bug, and diagnosing it is the practical-notes section below; the short version is that our default inverse-Wishart prior cannot believe in zero variance. File it, and first collect the payoff the chapter exists for.

## The Payoff: Individual-Level Posteriors

The sampler's `B_draws` array holds something neither @sec-mixed-msl nor any amount of MSL effort could produce: *joint posterior draws of every respondent's coefficient vector*, with population-level uncertainty propagated through automatically --- $\bar{\bfbeta}$ and $\Sigma$ wiggled underneath every person's draws, the one-step architecture @sec-limits demanded. Person-level point estimates and honest intervals are now sample statistics:

```{r}
nd <- dim(fit$B_draws)[1]                   # pool the two chains' person-draws
Bd <- array(NA_real_, c(2*nd, n_id, K))     # (rbind would flatten a 3-D array)
Bd[1:nd, , ] <- fit$B_draws
Bd[(nd+1):(2*nd), , ] <- fit2$B_draws
bhat_price <- colMeans(Bd[, , 6])           # posterior mean per person
blo <- apply(Bd[, , 6], 2, quantile, 0.05)
bhi <- apply(Bd[, , 6], 2, quantile, 0.95)
```

```{r}
#| label: fig-hb-individuals
#| fig-cap: "The deliverable: posterior means (dots) and 90% credible intervals (segments) of the price coefficient for 40 randomly chosen respondents, ordered by posterior mean, with each respondent's true coefficient marked (x). Intervals are person-specific, honest, and--for most respondents--cover the truth."
#| fig-width: 6.5
#| fig-height: 4
set.seed(171)
show <- sample(n_id, 40)
o <- show[order(bhat_price[show])]
dfp <- data.frame(rank = 1:40, mean = bhat_price[o], lo = blo[o], hi = bhi[o],
                  truth = study$betas_true[o, 6])
ggplot(dfp, aes(x = rank)) +
    geom_segment(aes(xend = rank, y = lo, yend = hi), color = "grey60") +
    geom_point(aes(y = mean), size = 1.2) +
    geom_point(aes(y = truth), shape = 4, size = 1.8) +
    labs(x = "respondent (ordered)", y = expression(beta[price]))
```

```{r}
c(cor_with_truth = round(cor(bhat_price, study$betas_true[, 6]), 2),
  coverage_90 = round(mean(study$betas_true[, 6] >= blo &
                           study$betas_true[, 6] <= bhi), 2))
```

Person-by-person tracking of the truth on par with @sec-limits's conditional means[^2] --- but now each respondent carries an interval, and those intervals *cover at close to their nominal rate*. That coverage number is quietly the chapter's most important line: it certifies that the person-level uncertainty statements mean what they say, the exact property @sec-limits proved the two-step likelihood route could not deliver. Shrinkage is visible too --- posterior means span a narrower range than the truths, for @sec-gibbs's precision-weighted reasons --- so the @sec-limits warning stands: the spread of *tastes* is read from $\Sigma$'s posterior, never from the spread of person-level means.

[^2]: Slightly lower correlation than @sec-limits's conditional means achieve on the same data, in fact --- and the reason is instructive: the HB model here estimates all six coefficients as random with free covariance, spreading eight tasks of personal information across a richer parameterization than the (conveniently truth-matching) three-coefficient model @sec-limits inherited. Generality has a price; honesty about not knowing the truth's structure is what the extra 0.05 of correlation bought.

## HB and MSL on the Same Data

The comparison @sec-msl bookmarked. Same dataset, two engines:

```{r}
msl <- study$theta_hat      # @sec-mixed-msl: means (6) + log-sds (3 random)
comp <- data.frame(
    truth = c(study$bar_true, study$sd_true[study$rc]),
    msl   = c(msl[1:6], exp(msl[7:9])),
    hb    = c(colMeans(bar_draws), colMeans(sd_draws)[study$rc])
)
rownames(comp) <- c(colnames(X), paste0("sd_", colnames(X)[study$rc]))
round(comp, 2)
```

Substantive agreement on every population mean and on the heterogeneity ordering, with the HB numbers running slightly larger --- partly the IW floor (below), partly the specification gap (HB's phantom heterogeneity on the fixed coefficients absorbs scale). The fair scorecard, then [and it matches the literature's: @train2001; @benakiva2019, Ch. 7]:

- **Population parameters, modest spec:** a tie --- both engines recover them, in comparable time here.
- **Full covariance matrices:** HB, walking away. MSL's 27-parameter Cholesky climb is HB's one extra `rinvwishart()` line.
- **Individual-level inference:** HB by design --- native draws with honest coverage versus a bolt-on with a plug-in.
- **Sharp-null economy:** MSL retains one genuine edge --- when you *want* a coefficient exactly fixed (no heterogeneity), you just don't randomize it, while the IW-equipped hierarchy resists zeros (next section) and constraining it takes work.
- **Tuning burdens:** a wash, honestly --- MSL trades draw counts and starting values for HB's burn-in judgment and step adaptation. Neither is push-button; both are manageable with the diagnostics this book has drilled.

## Validation Against `bayesm`

The reference implementation is `rhierMnlRwMixture` [@rossi2005] --- the same model with a mixture-of-normals population layer, set to one component to match @eq-hb-mnl, and the same default priors. Its data format wants each person's tasks as a list of per-occasion design matrices:

```{r}
#| cache: false
#| message: false
#| results: hide
library(bayesm)
J <- 3; T <- 8
lgtdata <- lapply(1:n_id, function(n) {
    rows <- which(id == n)
    Xn <- X[rows, ]
    list(y = max.col(t(matrix(y[rows], nrow = J))),
         X = lapply(1:T, function(t) Xn[((t-1)*J + 1):(t*J), , drop = FALSE]))
})
set.seed(172)
outb <- rhierMnlRwMixture(Data = list(p = J, lgtdata = lgtdata),
                          Prior = list(ncomp = 1),
                          Mcmc = list(R = 5000, keep = 2, nprint = 0))
```

```{r}
keepb <- 1001:2500     # kept draws: keep = 2 stores every 2nd, so this discards 2,000 iterations as burn-in
mu_b <- t(sapply(keepb, function(s) outb$nmix$compdraw[[s]][[1]]$mu))
sd_b <- t(sapply(keepb, function(s) {
    rooti <- outb$nmix$compdraw[[s]][[1]]$rooti
    sqrt(diag(chol2inv(t(rooti))))
}))
round(rbind(ours_bar = colMeans(bar_draws), bayesm_bar = colMeans(mu_b),
            ours_sd  = colMeans(sd_draws),  bayesm_sd  = colMeans(sd_b)), 2)
```

Two independently coded samplers --- ours, sixty transparent lines; theirs, a compiled industry standard --- agreeing on both layers of the posterior.[^3] Note that `bayesm` shows the *same* elevated standard deviations on the truly-fixed coefficients: the phenomenon is the prior's, not our code's, which is exactly what the next section establishes on purpose.

[^3]: To numerical-comparison precision for MCMC: posterior means from finite chains carry Monte Carlo error, so "agreement" means within a couple of MC standard errors, not five decimals. The @sec-mnl-mle five-decimal standard was for deterministic optimizers.

## Practical Notes

**The inverse-Wishart floor --- priors on $\Sigma$ matter.** Our recovery table showed $\hat\sigma \approx 0.6$ for coefficients whose true heterogeneity is zero. The mechanism: the IW prior with scale $V_0 = \nu_0 I$ puts essentially no mass on variances near zero --- it is a prior *against* homogeneity --- and eight tasks per person cannot overrule it (a person's eight choices are consistent with modest taste variation plus Gumbel noise). Watch the prior's hand move, by re-running with the scale matrix shrunk an order of magnitude:

```{r}
#| cache: false
set.seed(173)
fit_tight <- hb_mnl(y, X, task_id, id, S_iter = 4000, burn = 1000,
                    V0 = diag(K) * (K + 3) * 0.1)
round(rbind(default_V0 = colMeans(sd_draws),
            tight_V0   = colMeans(fit_tight$sd[-(1:1000), ]),
            truth      = study$sd_true), 2)
```

The phantom standard deviations drop by nearly half --- and the real ones shrink too, `sd_price` now *under* the truth. Neither prior is "right": the IW simply cannot express "some variances near zero, others large" --- its single scale parameter couples them --- and this is a known, documented limitation with active remedies in the literature (separation strategies; tightened scales chosen by prior predictive checks, as in @allenby2014). The working discipline is the one @sec-bayes-mnl taught: *run the sensitivity, report it, and interpret heterogeneity magnitudes with the prior in view.* With hundreds of tasks per person the data would overwhelm all of this; with eight, the prior is a coauthor, and pretending otherwise is the error.

**Sign constraints via transformation.** Our population layer lets some simulated consumers like high prices (@sec-mixed-msl's blemish, inherited). The standard repair reparameterizes: model $\theta_n$ with $\beta_{n,\text{price}} = -e^{\theta_n}$ --- @sec-draws's lognormal transformation --- so every person's price coefficient is negative while $\theta_n$ enjoys the normal hierarchy untouched. In the sampler, exactly one line changes: `persll()`'s utility computation applies the transform before multiplying. The blocks, the conjugacy, the adaptation --- all survive, because the hierarchy never cared what the coefficients *mean*. We defer the demonstration to @sec-extensions, where transformed coefficients earn a starring role in the budget-constrained model.

**Reporting: two different "heterogeneity" statements.** The posterior of $\bar{\bfbeta}$ answers "where is the average consumer" --- and its uncertainty shrinks with $N$. The posterior of the *taste distribution* $N(\bar{\bfbeta}, \Sigma)$ answers "what consumers are out there" --- and its spread does not shrink with $N$; it is an estimate of real variety. Deliverables built on the wrong one (a "share who would pay X" computed from $\bar{\bfbeta}$'s posterior instead of the taste distribution) are a classic applied error [@chapman2013's "averages hide the distribution," now with the machinery to do it right]. @sec-postestimation builds the correct versions.

One save for the chapters ahead --- the pooled draws, which @sec-postestimation will spend freely:

```{r}
saveRDS(list(bar_draws = bar_draws, sd_draws = sd_draws, B_draws = Bd,
             X_names = colnames(X)),
        file = "../data/hb_mnl_fit.rds")
```

## Key Learnings

- The HB-MNL is the mixed MNL wearing priors (@eq-hb-mnl), and its sampler is pure assembly: an all-persons vectorized Metropolis step on the @sec-mnl-mle kernel (adaptive during burn-in, ~30% acceptance), a conjugate normal draw for $\bar{\bfbeta}$, an inverse-Wishart draw for $\Sigma$ --- Metropolis-within-Gibbs as @sec-gibbs designed it, in sixty lines that run two chains in minutes.
- The full covariance matrix that strained MSL is free here --- no draws, no Cholesky optimization, no $R$ to choose. The engines agree on population parameters where their specifications overlap (and `bayesm` agrees with our sixty lines on everything).
- The payoff is @fig-hb-individuals: every respondent gets a posterior --- point, interval, the works --- with population uncertainty propagated in one step, and the 90% intervals *actually cover* ~90% of the true individual coefficients. This is the deliverable @sec-limits proved likelihood could not honestly produce.
- Shrinkage discipline carries over: person means understate taste spread by construction; read heterogeneity from $\Sigma$, and never confuse the posterior of the average consumer with the distribution of consumers.
- The priors are coauthors at $T = 8$: the default IW cannot believe in zero variance (phantom sds near 0.6 on truly-fixed coefficients, reproduced by `bayesm`), and tightening $V_0$ moves every variance --- run the sensitivity, report it, interpret accordingly.
- Sign constraints are one transformed line away (lognormal price), because the hierarchy is agnostic about coefficient meaning --- the extensibility that @sec-extensions will exploit.
