::: {.content-hidden}
{{< include ../latex-header.tex >}}
:::

# Post-Estimation: Inference, WTP, and Market Simulation {#sec-postestimation}

```{r}
#| label: setup
#| message: false
#| echo: false
library(ggplot2)
theme_set(theme_minimal())
```

No one who commissions a choice study wants a coefficient. They want answers: *what is the memory upgrade worth in dollars? What share does our configuration win at \$1,499? Who are the price-insensitive customers?* Between the estimates of Parts II--IV and those answers sits one more layer of craft --- transforming parameters into decision quantities, with uncertainty carried through honestly. The layer is thin (mostly arithmetic on things we already have) but it is where analyses most often go wrong in practice, because the arithmetic has traps with famous names. This chapter builds the transformations and disarms the traps.

Two fits power the chapter, both already on disk: the plain MNL of @sec-mnl-mle (refit inline below --- it takes a second) and the HB-MNL draws saved by @sec-hb-mnl.

```{r}
study   <- readRDS("../data/laptop_study.rds")
laptops <- study$data
X <- model.matrix(~ brand + ram + screen + price, data = laptops)[, -1]
y <- laptops$choice
task_id <- cumsum(!duplicated(laptops[, c("id", "task")]))
K <- ncol(X)

loglik_mnl <- function(beta, y, X, task_id) {
    V <- as.vector(X %*% beta); Vmax <- ave(V, task_id, FUN = max)
    eV <- exp(V - Vmax)
    sum((V - Vmax - log(rowsum(eV, task_id)[task_id]))[y == 1])
}
mle <- optim(rep(0, K), loglik_mnl, y = y, X = X, task_id = task_id,
             method = "BFGS", control = list(fnscale = -1), hessian = TRUE)
vc  <- solve(-mle$hessian)

hb <- readRDS("../data/hb_mnl_fit.rds")     # bar_draws, sd_draws, B_draws
```

## Inference on Functions of Parameters

The recurring problem: the quantity of interest is $g(\bfbeta)$, not $\bfbeta$ --- a ratio, a probability, a share --- and it needs a standard error or interval. Three standard tools, in ascending order of generality.

**The delta method** linearizes: if $\hat{\bfbeta} \sim N(\bfbeta, V)$ approximately, then

$$
g(\hat{\bfbeta}) \;\dot\sim\; N\!\left( g(\bfbeta), \; \nabla g' \, V \, \nabla g \right)
$$ {#eq-delta}

--- the function's gradient maps parameter uncertainty into quantity uncertainty. Exact for linear $g$, an approximation otherwise, and increasingly strained as $g$ curves (ratios near a small denominator being the canonical strain). We used a special case in @sec-mixed-msl to convert log-scale standard errors; here is the general workhorse. The one-line reason: a first-order expansion $g(\hat{\bfbeta}) \approx g(\bfbeta) + \nabla g(\bfbeta)'(\hat{\bfbeta} - \bfbeta)$ makes $g(\hat{\bfbeta})$ approximately a linear function of an approximately normal vector, and the variance of a linear function $\mathbf{a}'\hat{\bfbeta}$ is $\mathbf{a}' V \mathbf{a}$.

**Krinsky--Robb** [@krinsky1986] simulates instead: draw many $\bfbeta^{(r)} \sim N(\hat{\bfbeta}, V)$ from the estimated asymptotic distribution, push each through $g$, and read summaries off the resulting sample. No gradients, no linearization --- @sec-draws's summarize-by-sampling logic applied to frequentist output. It is the poor astronomer's posterior: same mechanics as the Bayesian version, different justification.

**Bayesian draws** make the problem dissolve: the posterior sample of $\bfbeta$ *is* a posterior sample of $g(\bfbeta)$ after applying $g$ row by row --- the quiet superpower advertised in @sec-bayes-mnl, no approximation anywhere.

Watch all three on one question --- *the dollar value of upgrading 16GB to 32GB* --- and note how they cluster:

$$
g(\bfbeta) = \frac{\beta_{\texttt{ram32}} - \beta_{\texttt{ram16}}}{-\beta_{\texttt{price}}} \times \$1{,}000
$$ {#eq-wtp-ratio}

(Utility difference of the upgrade divided by the utility of a dollar: the scale factor from @sec-mnl-identification cancels in the ratio, which is precisely why ratios are the preferred currency.)

```{r}
g <- function(b) 1000 * (b[4] - b[3]) / (-b[6])

## delta method: numerical gradient of g (via @sec-optimization's num_grad idea)
grad_g <- sapply(1:K, function(k) {
    e <- replace(numeric(K), k, 1e-6)
    (g(mle$par + e) - g(mle$par - e)) / 2e-6
})
se_delta <- sqrt(t(grad_g) %*% vc %*% grad_g)

## Krinsky-Robb: draws from the asymptotic normal (@sec-mvnorm transform)
set.seed(210)
kr <- apply(mle$par + t(chol(vc)) %*% matrix(rnorm(K*5000), K), 2, g)

## Bayesian: transform the posterior draws of the population means
bayes <- apply(hb$bar_draws, 1, g)

round(data.frame(
    method   = c("delta", "Krinsky-Robb", "Bayes (HB bar)"),
    estimate = c(g(mle$par), mean(kr), mean(bayes)),
    se_or_sd = c(se_delta, sd(kr), sd(bayes)),
    lo95     = c(g(mle$par) - 1.96*se_delta, quantile(kr, .025), quantile(bayes, .025)),
    hi95     = c(g(mle$par) + 1.96*se_delta, quantile(kr, .975), quantile(bayes, .975))
)[, -1], 0)
```

Same answer to rounding --- roughly \$250 with an interval of $\pm$\$70-ish --- with the small differences instructive: delta and Krinsky--Robb share inputs and differ only by linearization; the Bayesian row differs slightly more because it comes from a different model fit (the HB's population mean) on different data-processing of the same study. When the three *disagree materially*, the usual culprit is a denominator too close to zero for @eq-delta's linearization --- at which point trust the sampling-based rows.

## Willingness-to-Pay, and Its Traps

@eq-wtp-ratio is a **marginal WTP**: the rate at which a consumer trades dollars for an attribute, all else equal. For small, single-attribute comparisons it is the right object. Two traps surround it.

**Trap one: the heterogeneous ratio.** With HB output it is tempting to compute each *person's* WTP by applying @eq-wtp-ratio to their $\bfbeta_n$ draws. Legitimate --- and dangerous, because our population model makes $\beta_{\text{price},n}$ normal, so some mass sits near (and past) zero, and dividing by near-zero produces WTP draws in the tens of thousands of dollars with the wrong sign. Watch:

```{r}
wtp_n <- 1000 * (hb$B_draws[, , 4] - hb$B_draws[, , 3]) / (-hb$B_draws[, , 6])
round(c(mean = mean(wtp_n), median = median(wtp_n),
        q01 = quantile(wtp_n, .01), q99 = quantile(wtp_n, .99)), 0)
```

The mean is inflated (or even nonsensical) while the median and inner quantiles are sane --- the textbook signature of a ratio with a denominator crossing zero. Remedies, in rising order of principle: report medians and quantiles, never means; re-estimate with a sign-constrained (lognormal) price coefficient, as flagged in @sec-hb-mnl; or reparameterize the whole model in *WTP space* so the ratio is a parameter rather than a quotient [@train2016 and @benakiva2019 discuss both]. What is not acceptable is publishing the mean of an unconstrained ratio --- a mistake with a long paper trail.

**Trap two: pseudo-WTP versus market value.** The ratio answers "utility compensation for one attribute, one option, no competition." It does *not* answer "what is the feature worth to a firm" --- in a market, a better product steals share from rivals, rivals reprice, and the value of a feature is an equilibrium object. @allenby2014 make the accounting vivid: ratio-based "WTP" can overstate what pricing can capture by an order of magnitude or more, and the honest firm-side calculation runs through the market simulator below (and, fully, through equilibrium pricing --- beyond our scope but built from exactly these pieces). The economist's consumer-surplus version for the logit has a closed form worth knowing. The expected maximum utility of a Gumbel choice set is $\ln \sum_j e^{V_j}$ plus a constant (Euler's $\gamma \approx 0.577$, the Gumbel mean --- it cancels the moment we take *differences* between scenarios), a classic result we state rather than derive [@small1981; @train2009, Section 3.5] --- @sec-mnl's log-sum, in its third and final starring role. Dividing the change in expected utility by $-\beta_{\text{price}}$, the marginal utility of a dollar, converts utils to dollars --- legitimate because our linear-in-price utility makes that marginal utility constant. The *compensating variation* for any change in the choice set is therefore:

$$
CV = \frac{ \ln \sum_j e^{V_j^{\text{after}}} - \ln \sum_j e^{V_j^{\text{before}}} }{ -\beta_{\text{price}} }
$$ {#eq-logsum-cv}

```{r}
## market context: three laptops; upgrade the Acer's memory 16 -> 32
V_of <- function(prods, b) as.vector(prods %*% b)
market <- rbind(acer  = c(0, 0, 1, 0, 1, 1.2),     # dell,apple,ram16,ram32,screen15,price
                dell  = c(1, 0, 0, 1, 1, 1.6),
                apple = c(0, 1, 0, 1, 0, 2.0))
market2 <- market; market2["acer", 3:4] <- c(0, 1)

cv <- 1000 * (log(sum(exp(V_of(market2, mle$par)))) -
              log(sum(exp(V_of(market,  mle$par))))) / (-mle$par[6])
round(c(ratio_wtp = g(mle$par), market_cv = cv), 0)
```

The market-context value is a fraction of the all-else-equal ratio --- because in a market with good substitutes, consumers escape a mediocre product rather than pay for its upgrade. Neither number is wrong; they answer different questions, and attaching the right number to the right question is the analyst's job.

## Market Simulation

The workhorse deliverable. A **market simulator** takes a hypothetical market --- a set of products with attributes --- and returns predicted choice shares, built from nothing but the model's probability formula and the estimated (distribution of) coefficients. With HB draws, shares come with uncertainty attached: compute the share under each posterior draw of each person's $\bfbeta_n$, average over people (the market), and summarize over draws (the uncertainty):

```{r}
simulate_shares <- function(products, B_draws) {
    S <- dim(B_draws)[1]; Jm <- nrow(products)
    shares <- matrix(NA_real_, S, Jm)
    for (s in 1:S) {
        V <- B_draws[s, , ] %*% t(products)      # persons x products
        V <- V - apply(V, 1, max)                 # @sec-logsumexp, always
        P <- exp(V) / rowSums(exp(V))
        shares[s, ] <- colMeans(P)                # average person = market share
    }
    colnames(shares) <- rownames(products)
    shares
}

sh <- simulate_shares(market, hb$B_draws)
round(rbind(share = colMeans(sh),
            lo95 = apply(sh, 2, quantile, .025),
            hi95 = apply(sh, 2, quantile, .975)), 3)
```

Read the structure of `simulate_shares()` once more, because it encodes the two averaging levels that must not be confused: `colMeans(P)` averages *people* (heterogeneity --- part of the market, inside each draw), while summaries over `s` average *posterior draws* (uncertainty --- reported as intervals). Collapse them in the wrong order --- computing shares at the *average* $\bfbeta$ --- and the simulator inherits the MNL's IIA at the population level, forfeiting exactly the substitution realism the heterogeneity bought (@sec-mixed-msl's aggregation insight, now with money on the line).

Demand curves are the simulator in a loop --- sweep one product's price, watch its share:

```{r}
#| label: fig-demand-curve
#| fig-cap: "Simulated demand curve for the Acer configuration against its two rivals, with 95% posterior bands: the market simulator swept over price. The curve's every point carries honest uncertainty inherited from the full posterior."
#| fig-width: 6
#| fig-height: 3.4
grid <- seq(0.8, 2.4, by = 0.1)
dc <- sapply(grid, function(p) {
    m <- market; m["acer", 6] <- p
    simulate_shares(m, hb$B_draws)[, "acer"]
})
dfd <- data.frame(price = grid, share = colMeans(dc),
                  lo = apply(dc, 2, quantile, .025),
                  hi = apply(dc, 2, quantile, .975))
ggplot(dfd, aes(price, share)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey80") +
    geom_line() +
    labs(x = "Acer price ($1,000s)", y = "preference share")
```

One vocabulary discipline, due to @chapman2013 and hard-won industry experience: the simulator outputs **preference shares** --- shares of choices among these products, by these respondents, under forced attention to the choice task. Real *market* shares also run through awareness, distribution, availability, and the option to buy nothing. Preference shares inform decisions powerfully (especially *differences* between scenarios, where the omitted factors partially cancel); presenting them as market share forecasts is the fastest way for a good model to lose a client's trust. (The "buy nothing" piece, at least, the model can absorb --- the no-choice option of @sec-extensions.)

## Elasticities and Substitution

Differentiating the MNL probability gives the textbook elasticity formulas --- own-price $e_{jj} = \beta_p p_j (1 - P_j)$ and the IIA-flavored cross-price elasticity that is *identical for all rivals* [@train2009, Section 3.6]. Rather than memorize formulas, note that the simulator computes any elasticity numerically --- nudge a price, difference the shares --- and, run on HB draws, it exhibits the realistic substitution the population mixture creates. The @sec-iia experiment, replayed with estimated parameters:

```{r}
bump <- market; bump["dell", 6] <- bump["dell", 6] * 0.9   # Dell cuts price 10%
delta <- colMeans(simulate_shares(bump, hb$B_draws)) - colMeans(sh)
round(rbind(base_share = colMeans(sh), change = delta,
            pct_change = delta / colMeans(sh)), 3)
```

Dell's gain comes disproportionately from the rival whose *buyers look most like Dell's buyers* in taste space --- proportional-substitution arithmetic broken by heterogeneity alone, no nests required. This table, produced for a client's actual scenario, is the applied payoff of every substitution-pattern discussion since @sec-mnl.

## Using the Individual-Level Estimates

Finally, the deliverable that only Part III could license: person-level quantities. Each respondent's posterior WTP (medians, per the trap above) supports targeting --- and one summary picture enforces this book's most repeated interpretive warning:

```{r}
#| label: fig-wtp-heterogeneity
#| fig-cap: "The distribution of respondent-level median WTP for the 32GB upgrade. The dashed line is the population average -- a number that describes almost nobody: most respondents sit well below it, a long tail sits far above. Averages hide the distribution; the distribution is the market."
#| fig-width: 6
#| fig-height: 3.2
wtp_person <- apply(wtp_n, 2, median)          # per-person posterior median
ggplot(data.frame(w = wtp_person), aes(w)) +
    geom_histogram(bins = 50, fill = "grey70") +
    geom_vline(xintercept = mean(wtp_person), linetype = "dashed") +
    coord_cartesian(xlim = quantile(wtp_person, c(.005, .995))) +
    labs(x = "median WTP for 32GB upgrade ($)", y = "respondents")
```

Segments fall out of such distributions directly (the right tail is the premium-upgrade target list, by respondent ID), and every claim about them inherits posterior uncertainty from the draws. This is @sec-limits's wish list, fully cashed: person-level, honest, decision-ready.

## Key Learnings

- Decision quantities are functions of parameters, and their uncertainty comes from three tools: the delta method (@eq-delta; fast, linearized), Krinsky--Robb (simulate the asymptotic normal), and posterior draws (exact, trivial). They agree when $g$ is tame; when they diverge, suspect the linearization.
- Marginal WTP is a scale-free ratio (@eq-wtp-ratio) --- but ratios of heterogeneous coefficients explode when the denominator crosses zero (report medians; or constrain signs; or move to WTP-space), and ratio-WTP is *not* market value: the log-sum compensating variation (@eq-logsum-cv) and the simulator answer the market question, usually with much smaller numbers [@allenby2014].
- The market simulator is the model's probability formula pointed at hypothetical products; with HB draws it averages people *inside* each draw and summarizes uncertainty *across* draws --- never share-at-the-average-beta, which resurrects IIA.
- Simulated shares are preference shares, not market shares (@chapman2013); differences between scenarios travel better than levels.
- Heterogeneity is the deliverable: person-level WTP distributions, targeting lists, and substitution patterns that put competition where the taste overlap is. The average consumer, meanwhile, describes almost nobody --- plot the distribution.
