::: {.content-hidden}
{{< include ../latex-header.tex >}}
:::

# Extensions: Building Your Own Model {#sec-extensions}

```{r}
#| label: setup
#| message: false
#| echo: false
library(ggplot2)
theme_set(theme_minimal())
```

Every chapter until now taught you models other people built. This one teaches the move that matters for your own research: *changing* a model --- because your data have a feature the standard models ignore, or your theory implies behavior they cannot express. The introduction promised that the true value of choice modeling is the ability to create tailor-made models; here is where the promise is kept, three times, on three extensions drawn from live research literatures.

The method is the one this book has been rehearsing since @sec-choice, now stated as a recipe:

1. **Change the model** --- a new term in utility, a new choice rule $h$, a new distribution somewhere in the hierarchy.
2. **Re-derive the choice probability** (or recognize that a known one still applies).
3. **Write the simulator first.** If you cannot simulate it, you do not understand it yet.
4. **Modify the likelihood or the sampler** --- usually a handful of lines in code you already have.
5. **Simulate-and-recover before real data ever enters.** The answer key is the only proof your estimator works.

Watch the recipe's economy in what follows: each extension changes remarkably little code, because Part I's architecture --- stacked data, kernel likelihood functions, modular samplers --- was designed to be extended. That was the point all along.

## Extension 1: The No-Choice Option {#sec-ext-nochoice}

**The problem.** Our laptop study forces a choice: every task ends with one of three laptops "bought." Real shoppers walk away, and a model estimated on forced choices systematically overstates demand --- every simulated market in @sec-postestimation implicitly assumed everyone buys *something*. The standard repair adds a **no-choice option** ("I would not buy any of these") to each task [@brazell2006].

**The model change.** The none option is a fourth alternative whose utility has no attributes --- nothing about it varies --- so it is pure constant plus noise:

$$
U_{nt,\text{none}} = \gamma + \varepsilon_{nt,\text{none}}
$$ {#eq-none-utility}

One new parameter. And here @sec-data-intercept's groundwork pays off: in the long data format, the none option is *just another row* --- attribute columns zero, a `none` dummy equal to one --- and the MNL machinery does not know or care that one alternative is special. The choice probabilities, likelihood, gradient, and code are all unchanged.[^1]

[^1]: The identification bookkeeping: adding a constant-utility alternative is what *anchors* the utility scale for "buying nothing," making statements like "share who buy nothing at these prices" meaningful --- the quantity @sec-postestimation flagged as missing. The deeper behavioral question --- is the none utility really constant across tasks? --- is live research: @hung2025nochoice find respondents update the none option's appeal as tasks teach them market prices, and model exactly that.

**Simulate, estimate, recover** --- the recipe executed in one block. The simulator is @sec-sim-mnl's with a fourth row per task; the estimator is *literally* `loglik_mnl()`:

```{r}
#| cache: false
set.seed(220)
sim_none_data <- function(N, T, beta, gamma_none) {
    J <- 3; n_rows <- N * T * J
    df <- data.frame(
        id = rep(seq_len(N), each = T*J),
        task = rep(rep(seq_len(T), each = J), times = N),
        alt = rep(1:J, times = N*T),
        brand = factor(sample(c("Acer","Dell","Apple"), n_rows, TRUE),
                       levels = c("Acer","Dell","Apple")),
        ram = factor(sample(c("8","16","32"), n_rows, TRUE),
                     levels = c("8","16","32")),
        screen = factor(sample(c("13","15"), n_rows, TRUE),
                        levels = c("13","15")),
        price = round(runif(n_rows, 0.8, 2.4), 2))
    X <- cbind(model.matrix(~ brand + ram + screen + price, df)[, -1],
               none = 0)
    ## append one none row per task: all-zero attributes, none dummy = 1
    Xn <- X[df$alt == 1, ]; Xn[] <- 0; Xn[, "none"] <- 1
    X  <- rbind(X, Xn)
    ids  <- c(df$id,   df$id[df$alt == 1])
    tsks <- c(df$task, df$task[df$alt == 1])
    ord  <- order(ids, tsks, c(df$alt, rep(4, sum(df$alt == 1))))
    X <- X[ord, ]; ids <- ids[ord]; tsks <- tsks[ord]

    task_id <- cumsum(!duplicated(cbind(ids, tsks)))
    U <- as.vector(X %*% c(beta, gamma_none)) - log(-log(runif(nrow(X))))
    choice <- as.integer(ave(U, task_id, FUN = function(u) u == max(u)))
    list(X = X, y = choice, task_id = task_id, id = ids)
}

beta_true <- c(dell = 0.5, apple = 1.0, ram16 = 0.6, ram32 = 0.9,
               screen15 = 0.3, price = -1.2)
gamma_true <- -0.5                      # none is mildly unattractive
d1 <- sim_none_data(500, 8, beta_true, gamma_true)
mean(rowsum(d1$y * (d1$X[, "none"] == 1), d1$task_id))   # share choosing none
```

About a fifth of tasks end in "none" --- real demand information the forced design threw away. Estimation requires nothing new:

```{r}
#| cache: false
loglik_mnl <- function(beta, y, X, task_id) {           # @sec-mnl-mle, untouched
    V <- as.vector(X %*% beta); Vmax <- ave(V, task_id, FUN = max)
    eV <- exp(V - Vmax)
    sum((V - Vmax - log(rowsum(eV, task_id)[task_id]))[y == 1])
}
fit1 <- optim(rep(0, 7), loglik_mnl, y = d1$y, X = d1$X, task_id = d1$task_id,
              method = "BFGS", control = list(fnscale = -1), hessian = TRUE)
round(rbind(truth = c(beta_true, none = gamma_true),
            est = fit1$par,
            se = sqrt(diag(solve(-fit1$hessian)))), 2)
```

Recovery, including the new parameter. Total new code: a simulator variant. Total new estimation code: zero --- the none option was a *data* change, which is the least expensive kind of extension there is.

**The dual-response refinement**, in brief, because practice uses it heavily [@diener2006; @brazell2006]: asking "which would you choose?" *and then* "would you actually buy it?" keeps the forced task's statistical efficiency while measuring purchase intent. The likelihood treats the two answers as two choices sharing parameters --- the first over the $J$ products, the second between the winner and none --- and @diener2006's practical insight is that the second response can be *stacked as an additional pseudo-task* in the long data, after which standard MNL/HB software estimates it unmodified: another extension executed entirely in the data layer. The modern frontier connects the none utility to economic primitives --- @hung2025dual derive it from a budget constraint, which is precisely where our second extension goes.

## Extension 2: A Budget Constraint {#sec-ext-budget}

**The problem.** For laptops --- or cars, or houses --- some shoppers simply *cannot* pay premium prices, no matter their tastes. A linear price term says a \$2,400 laptop is merely less attractive; a **budget** says it is not an option. Models that ignore binding budgets misestimate price response exactly where pricing decisions live, overstating premium demand --- the bias @pachali2022 document in conjoint studies of exactly our product category.

**The model change** --- and this one, unlike Extension 1, changes the behavioral rule $h$ itself. Following the spirit of @pachali2022 (simplified to a common budget $\tau$ for transparency): alternatives priced above $\tau$ are *screened out* --- infeasible, never chosen --- and money left under the budget enters utility with diminishing returns:

$$
U_{ntj} = \begin{cases}
\x_{ntj}'\bfbeta + \alpha \log(\tau - p_{ntj}) + \varepsilon_{ntj} & p_{ntj} < \tau \\
-\infty & p_{ntj} \ge \tau
\end{cases}
$$ {#eq-budget-utility}

(The $\log(\tau - p)$ term is the utility of the outside numeraire --- what the unspent budget buys --- and it is where the price sensitivity now lives; there is no separate linear price coefficient.[^2]) The MNL probability survives with two amendments: the choice set shrinks to the feasible alternatives, and utilities gain curvature in price that grows explosive as $p \to \tau$. Both are *visible in behavior* --- shoppers never choosing above a ceiling, and price sensitivity concentrating near it --- which is what identifies $\tau$.

[^2]: This is the direct-utility formulation of @allenby2019 and @pachali2022: choose one product plus outside spending subject to the budget; substitute the constraint into utility; the $\log$ delivers diminishing marginal utility of money. Our simplification is the common $\tau$; the research version makes $\tau_n$ heterogeneous, which the road map below restores.

**Simulator first**, per the recipe --- and note how @eq-budget-utility transcribes:

```{r}
#| cache: false
set.seed(221)
sim_budget_data <- function(N, T, beta, alpha, tau) {
    J <- 3; n_rows <- N * T * J
    df <- data.frame(
        id = rep(seq_len(N), each = T*J),
        task = rep(rep(seq_len(T), each = J), times = N),
        alt = rep(1:J, times = N*T),
        brand = factor(sample(c("Acer","Dell","Apple"), n_rows, TRUE),
                       levels = c("Acer","Dell","Apple")),
        ram = factor(sample(c("8","16","32"), n_rows, TRUE),
                     levels = c("8","16","32")),
        screen = factor(sample(c("13","15"), n_rows, TRUE),
                        levels = c("13","15")),
        price = round(runif(n_rows, 0.8, 2.4), 2))
    X <- model.matrix(~ brand + ram + screen, data = df)[, -1]  # note: no price column
    V <- as.vector(X %*% beta) +
         ifelse(df$price < tau, alpha * log(tau - df$price), -Inf)
    U <- V - log(-log(runif(n_rows)))
    task_id <- cumsum(!duplicated(df[, c("id","task")]))
    df$choice <- as.integer(ave(U, task_id, FUN = function(u) u == max(u)))
    list(df = df, X = X, task_id = task_id)
}

beta_np <- c(dell = 0.5, apple = 1.0, ram16 = 0.6, ram32 = 0.9, screen15 = 0.3)
alpha_true <- 1.4; tau_true <- 2.6         # budget $2,600: binding for top prices
d2 <- sim_budget_data(500, 8, beta_np, alpha_true, tau_true)
max(d2$df$price[d2$df$choice == 1])         # nobody chose above tau... check
```

(A task where *all three* laptops exceed $\tau$ would have no feasible option; with our price range topping out at \$2,400 < $\tau$, the issue does not arise here, but your own version must decide what shoppers do then --- a none option, Extension 1, is the natural coupling. Model changes interact; simulators surface the interactions before referees do.)

**The likelihood** amends the @sec-mnl-mle kernel in exactly the two places the model changed --- the utility line gains the budget term, and infeasible rows carry $-\infty$ (which the log-sum-exp machinery handles gracefully: $e^{-\infty} = 0$ drops them from the denominator):

```{r}
#| cache: false
loglik_budget <- function(theta, y, X, price, task_id) {
    K <- ncol(X)
    beta <- theta[1:K]; alpha <- theta[K+1]; tau <- theta[K+2]
    if (tau <= max(price[y == 1])) return(-1e10)   # chosen>tau impossible
    V <- as.vector(X %*% beta) + alpha * log(pmax(tau - price, 1e-12))
    V[price >= tau] <- -Inf
    Vmax <- ave(V, task_id, FUN = max)
    eV <- exp(V - Vmax)
    sum((V - Vmax - log(rowsum(eV, task_id)[task_id]))[y == 1])
}
```

One wrinkle deserves its sentence: the feasibility screen makes the likelihood *zero* for any $\tau$ below the highest chosen price --- the data flatly refute such budgets --- so the parameter space has a data-dependent boundary, and the surface is non-smooth in $\tau$ at every observed price point. Textbook MLE regularity has left the building.[^3] Practical estimation is untroubled: profile the likelihood on a grid of $\tau$ (grid search's redemption arc, from @sec-grid-search) with the smooth parameters maximized at each point:

```{r}
#| cache: false
taus <- seq(2.45, 3.2, by = 0.025)
prof <- sapply(taus, function(tt) {
    f <- optim(c(rep(0, 5), 1), function(th, ...)
                   loglik_budget(c(th, tt), ...),
               y = d2$df$choice, X = d2$X, price = d2$df$price,
               task_id = d2$task_id,
               method = "BFGS", control = list(fnscale = -1, maxit = 300))
    c(tau = tt, ll = f$value, f$par)
})
best <- prof[, which.max(prof["ll", ])]
round(rbind(truth = c(beta_np, alpha = alpha_true, tau = tau_true),
            est   = c(best[3:7], best[8], best["tau"])), 2)
```

```{r}
#| label: fig-budget-profile
#| fig-cap: "The profile log-likelihood over the budget parameter. The data reject budgets below the highest chosen price outright (left boundary), climb to a well-defined peak near the true $2,600, and fall away as larger budgets flatten the log(tau - p) curvature the choices exhibit."
#| fig-width: 6
#| fig-height: 3
ggplot(data.frame(tau = prof["tau", ], ll = prof["ll", ]), aes(tau, ll)) +
    geom_line() + geom_point(size = 0.8) +
    geom_vline(xintercept = tau_true, linetype = "dashed") +
    labs(x = expression(tau~"($1,000s)"), y = "profile log-likelihood")
```

Recovery --- including a latent budget no survey question asked about, identified purely from where price sensitivity concentrates. **The research road map** from here is @pachali2022's: make $\tau_n$ person-specific with a hierarchical prior (the @sec-hb-mnl machinery --- the budget becomes one more transformed coefficient, log-normal to stay positive, sampled in the person-level Metropolis block), let stated budget questions enter as noisy *indicators* of $\tau_n$ rather than truth, and watch premium-price demand curves correct themselves by double-digit percentages. Every component is in your toolkit already.

## Extension 3: A Flexible Mixing Distribution {#sec-ext-flexmix}

**The problem.** Parts II--III assumed tastes are *normally* distributed --- always unimodal, symmetric, thin-tailed. Suppose the true market is polarized: a budget segment and a premium segment with little in between. A normal mixing distribution will report the average of two camps and the spread between them --- a distribution centered where almost nobody lives (@sec-postestimation's warning, at the population level).

**The model change**, following @train2016's *logit-mixed logit* in one dimension: replace the price coefficient's normal distribution with a **nonparametric distribution on a grid**. Fix $G$ candidate values $b_1, \ldots, b_G$ spanning the plausible range; estimate the probability mass $w_g$ on each, via a softmax in free parameters ($w_g \propto e^{\theta_g}$ --- @sec-latentclass's device). One normalization before coding it: adding a constant to every $\theta_g$ leaves the softmax weights unchanged --- the level unidentifiability of @sec-mnl-identification, back again --- so we pin $\theta_1 = 0$ and estimate the remaining $G-1$ free parameters. That is the `c(0, theta)` in the code below. The reader who has reached this chapter sees immediately what this *is*: a latent class model where the class locations are frozen to a grid and only the shares are estimated --- flexibility through many cheap classes rather than a few expensive ones.

**The estimation insight** that makes it practical --- and this is the trick worth stealing for other models: person $n$'s sequence likelihood at grid point $g$,
$$L_{ng} = \prod_{t=1}^{T} P_{nt}\big(y_{nt} \mid \beta_{\text{price}} = b_g\big)$$
--- the ordinary MNL likelihood of person $n$'s whole sequence with the price coefficient frozen at grid value $b_g$ (and the other coefficients at their step-1 values; `persll_at()` returns its log, one column of an $N \times G$ matrix per grid point) --- *does not depend on the weights*. Compute the $N \times G$ matrix of $L_{ng}$ **once**; thereafter the log-likelihood

$$
\ell(\bftheta) = \sum_n \log \sum_g w_g(\bftheta) \, L_{ng}
$$ {#eq-lml}

costs a matrix-vector product per evaluation --- the optimizer runs on a *precomputed* likelihood, and estimation takes seconds regardless of how expensive the kernel was [@train2016 reports 69 mixing parameters estimated in 18 seconds; ours is humbler]:

```{r}
#| cache: false
set.seed(222)
## simulate a polarized market: price coefficient is bimodal, others fixed
sim_laptop_design <- function(N, T, J) {
    n_rows <- N * T * J
    data.frame(id = rep(seq_len(N), each = T*J),
               task = rep(rep(seq_len(T), each = J), times = N),
               alt = rep(seq_len(J), times = N*T),
               brand = factor(sample(c("Acer","Dell","Apple"), n_rows, TRUE),
                              levels = c("Acer","Dell","Apple")),
               ram = factor(sample(c("8","16","32"), n_rows, TRUE),
                            levels = c("8","16","32")),
               screen = factor(sample(c("13","15"), n_rows, TRUE),
                               levels = c("13","15")),
               price = round(runif(n_rows, 0.8, 2.4), 2))
}
N <- 1000; T <- 8
df <- sim_laptop_design(N, T, 3)
X  <- model.matrix(~ brand + ram + screen + price, data = df)[, -1]
K  <- ncol(X)
task_id <- cumsum(!duplicated(df[, c("id","task")]))

b_fixed <- c(dell = 0.5, apple = 1.0, ram16 = 0.6, ram32 = 0.9, screen15 = 0.3)
bp_n <- ifelse(runif(N) < 0.55, rnorm(N, -2.0, 0.25), rnorm(N, -0.6, 0.25))
V <- as.vector(X[, 1:5] %*% b_fixed) + X[, 6] * bp_n[df$id]
U <- V - log(-log(runif(nrow(X))))
df$choice <- as.integer(ave(U, task_id, FUN = function(u) u == max(u)))
```

```{r}
#| cache: false
## step 1 (approximation, disclosed): fix non-price coefficients at MNL estimates
mnl <- optim(rep(0, K), loglik_mnl, y = df$choice, X = X, task_id = task_id,
             method = "BFGS", control = list(fnscale = -1))

## step 2: precompute person-by-gridpoint log-likelihoods
grid_b <- seq(-3.2, 0.2, length.out = 18)
id_of_task <- df$id[!duplicated(task_id)]
persll_at <- function(bp) {
    V <- as.vector(X[, 1:5] %*% mnl$par[1:5]) + X[, 6] * bp
    Vmax <- ave(V, task_id, FUN = max)
    eV <- exp(V - Vmax)
    logP <- (V - Vmax - log(rowsum(eV, task_id)[task_id]))[df$choice == 1]
    as.vector(rowsum(logP, id_of_task))
}
llng <- sapply(grid_b, persll_at)             # N x G, computed ONCE

## step 3: maximize over the weights -- each evaluation is trivial
loglik_lml <- function(theta, llng) {
    lw <- c(0, theta) - log(sum(exp(c(0, theta))))    # log softmax weights
    M  <- apply(sweep(llng, 2, lw, `+`), 1, max)
    sum(M + log(rowSums(exp(sweep(llng, 2, lw, `+`) - M))))
}
fit3 <- optim(rep(0, length(grid_b) - 1), loglik_lml, llng = llng,
              method = "BFGS", control = list(fnscale = -1, maxit = 1000))
w_hat <- exp(c(0, fit3$par)); w_hat <- w_hat / sum(w_hat)
```

```{r}
#| label: fig-lml
#| fig-cap: "The estimated nonparametric distribution of the price coefficient (bars: mass at each grid point) against the true bimodal density (line). Nobody told the model there were two segments; eighteen grid weights found them. A normal mixing distribution, by construction, could report only the average of the two camps."
#| fig-width: 6
#| fig-height: 3.2
true_dens <- function(x) 0.55*dnorm(x, -2.0, 0.25) + 0.45*dnorm(x, -0.6, 0.25)
gap <- diff(grid_b)[1]
ggplot(data.frame(b = grid_b, w = w_hat / gap), aes(b, w)) +
    geom_col(width = gap * 0.9, fill = "grey70") +
    geom_function(fun = true_dens, linewidth = 0.8) +
    labs(x = expression(beta[price]), y = "density")
```

Two modes, correct locations, correct proportions --- from choice data alone, with a mixing distribution the model was free to shape.[^4] The disclosed shortcut (fixing the other coefficients first) is the honest price of a demonstration; @train2016's full treatment estimates everything jointly, adds smoothness through spline or polynomial structure on the weights, and extends to multiple dimensions. And the Bayesian reader has already noticed that @eq-lml's grid weights would make a lovely Dirichlet-prior Gibbs block --- extensions compose across Parts II and III alike.

[^4]: Sharp eyes will connect this figure to @sec-hb-mnl's normal-prior HB: run on these polarized data, the normal hierarchy would shrink both camps toward a middle where no one lives. Flexible mixing is not decoration; on the wrong population shape, the normal assumption distorts every downstream deliverable in @sec-postestimation.

## Where to Go From Here

Three extensions, three different entry points into the machinery --- the data layer (none option), the behavioral rule (budget), the population distribution (flexible mixing) --- and none required more than a page of new code. The pattern generalizes to the research frontier: volumetric and multiple-discreteness models change the choice rule to Kuhn--Tucker conditions [@allenby2019]; WTP-space models reparameterize utility so preference heterogeneity lives in dollar units [@train2016; @benakiva2019]; context-dependent none utilities track what respondents learn across tasks [@hung2025nochoice]; and the unified framework of @marshall2002 shows ratings, rankings, and choices as different observation layers over one latent-utility core --- extensions of the *data* model rather than the utility model.

Your own extension will not be on this list; that is the point of it being yours. The recipe at the chapter's head does not care: change the model, re-derive the probability, *simulate first*, adjust the estimator, recover before real data. You now own every tool the recipe calls for --- which is what this book was for.

## Key Learnings

- Custom models are the recipe: model change $\to$ probability $\to$ **simulator first** $\to$ estimator tweak $\to$ simulate-and-recover. The architecture built in Part I is what makes each step small.
- The no-choice option is a *data-layer* extension: one all-zero row with an ASC per task, zero new estimation code, and it restores the walk-away margin that forced choice suppresses (with dual-response designs as the practice-standard refinement, also executable in the data layer).
- The budget constraint changes $h$ itself: screening plus $\log(\tau - p)$ curvature. Its likelihood has a data-dependent boundary and kinks in $\tau$ --- profiled grid search handles what BFGS regularity cannot --- and a latent budget is identified from behavior alone. The hierarchical version is @pachali2022 and everything it needs is in @sec-hb-mnl.
- Flexible mixing replaces the normal population with estimated grid weights --- latent class with frozen locations --- and Train's precomputed-likelihood trick ($L_{ng}$ once, then @eq-lml in microseconds) makes it nearly free. On polarized data it finds the two camps a normal distribution would average away.
- Extensions enter through the data, the rule, or the population --- and compose, with each other and with both estimation philosophies. Go build yours.
